---
title: "Exploring data sets for modeling"
output: 
  html_document:
    toc: true
    toc_float: true
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(here)
library(tidyverse)
library(sf)
source(here("R","functions-createVars.R"))
library(summarytools)
library(DataExplorer)
```

```{r}
IV <- readRDS(here("output","independentVars.rds")) %>% 
  st_as_sf()
DV <- readRDS(here("output","dependentVar.rds"))

nottypes <- c("wildlings","age","bio01","bio19","seeds.ExP","seeds.WALD","relelev")
```

Only plotting with randomseed == 1. Other randomseeds will have slightly different distributions of wildlings across height classes (but the same total number of wildlings).

```{r}
DV <- filter(DV, randomseed == 1)
```

```{r, include=FALSE}
theme_dotplot <- theme_bw(14) +
    theme(axis.text.y = element_text(size = rel(.75)),
        axis.ticks.y = element_blank(),
        axis.title.x = element_text(size = rel(.75)),
        axis.title.y = element_blank(),
        panel.grid.major.x = element_blank(),
        panel.grid.major.y = element_line(size = 0.5),
        panel.grid.minor.x = element_blank())
```

Also, only examining independent variables once --- not separately for each height class --- as they remain identical.

# Larix

```{r}
IVi <- filter(IV, species == "Larix") %>% 
  remove_zero_cols()
DVi <- filter(DV, species == "Larix")
dat <- DVi %>% 
  group_split(hgtclass) %>% 
  map(~ add_wildlings(IVi, .) %>% 
        st_drop_geometry())
```

'wildlings' below are those in height class 2.

```{r, echo = FALSE, results='asis'}
dfSummary(dat[[2]], plain.ascii = FALSE, style = "grid", 
          graph.magnif = 0.75, valid.col = FALSE, tmp.img.dir = "/tmp")
```

'wildlings' by height class and locality.

```{r}
bind_rows(dat, .id = "hgtclass") %>% 
  ggplot(aes(wildlings, fct_reorder(locality, wildlings, mean))) + 
         geom_point() + theme_dotplot + facet_grid(cols = vars(hgtclass))
ggplot(dat[[2]], aes(seeds.WALD, fct_reorder(locality, seeds.WALD, mean))) + geom_point() + theme_dotplot
ggplot(dat[[2]], aes(seeds.ExP, fct_reorder(locality, seeds.ExP, mean))) + geom_point() + theme_dotplot
```

```{r}
pairs(dat[[2]][sample(nrow(dat[[2]]), 1e4), nottypes])
```
 
# P.abies

```{r}
IVi <- filter(IV, species == "P.abies") %>% 
  remove_zero_cols()
DVi <- filter(DV, species == "P.abies")
dat <- DVi %>% 
  group_split(hgtclass) %>% 
  map(~ add_wildlings(IVi, .) %>% 
        st_drop_geometry())
```

'wildlings' below are those in height class 2.

```{r, echo = FALSE, results='asis'}
dfSummary(dat[[2]], plain.ascii = FALSE, style = "grid", 
          graph.magnif = 0.75, valid.col = FALSE, tmp.img.dir = "/tmp")
```

'wildlings' by height class and locality.

```{r}
bind_rows(dat, .id = "hgtclass") %>% 
  ggplot(aes(wildlings, fct_reorder(locality, wildlings, mean))) + 
         geom_point() + theme_dotplot + facet_grid(cols = vars(hgtclass))
ggplot(dat[[2]], aes(seeds.WALD, fct_reorder(locality, seeds.WALD, mean))) + geom_point() + theme_dotplot
ggplot(dat[[2]], aes(seeds.ExP, fct_reorder(locality, seeds.ExP, mean))) + geom_point() + theme_dotplot
```

```{r}
pairs(dat[[2]][sample(nrow(dat[[2]]), 1e4), nottypes])
```

# P.sitchensis-lutzii

```{r}
IVi <- filter(IV, species == "P.sitchensis-lutzii") %>% 
  remove_zero_cols()
DVi <- filter(DV, species == "P.sitchensis-lutzii")
dat <- DVi %>% 
  group_split(hgtclass) %>% 
  map(~ add_wildlings(IVi, .) %>% 
        st_drop_geometry())
```

'wildlings' below are those in height class 2.

```{r, echo = FALSE, results='asis'}
dfSummary(dat[[2]], plain.ascii = FALSE, style = "grid", 
          graph.magnif = 0.75, valid.col = FALSE, tmp.img.dir = "/tmp")
```

'wildlings' by height class and locality.

```{r}
bind_rows(dat, .id = "hgtclass") %>% 
  ggplot(aes(wildlings, fct_reorder(locality, wildlings, mean))) + 
         geom_point() + theme_dotplot + facet_grid(cols = vars(hgtclass))
ggplot(dat[[2]], aes(seeds.WALD, fct_reorder(locality, seeds.WALD, mean))) + geom_point() + theme_dotplot
ggplot(dat[[2]], aes(seeds.ExP, fct_reorder(locality, seeds.ExP, mean))) + geom_point() + theme_dotplot
```

```{r}
pairs(dat[[2]][sample(nrow(dat[[2]]), 1e4), nottypes])
```

```{r sessionInfo}
sessionInfo()
```