---
title: "Modeling relative establishment probability"
output: 
  html_document:
    toc: true
    toc_float: true
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, error=TRUE)
```

```{r}
library(here)
library(tidyverse)
source(here("R","functions.R"))
library(glmmTMB)
library(DHARMa)
library(furrr)
```

# Load data

```{r}
types <- read_csv(here('data','types.csv'))
contrasttype <- "T4"
nottypes <- c('species', 'locality','wildlings','age','bio01','bio19',
              'seeds.ExP','seeds.WALD','relelev','geometry','are')
```

```{r}
IV <- readRDS(here("output","independentVars.rds")) %>% 
  st_as_sf()
IV <- IV %>% 
  filter(`plantation forest` < 0.5) %>% 
  select(-`plantation forest`) %>% 
  rowwise() %>% 
  mutate(are = sum(c_across(any_of(types$type)))) %>% 
  filter(are >= 0.5) %>% 
  ungroup()
filter(IV, are > 1.01)
filter(IV, T4 > 0.5) %>%
  select(species, are, any_of(types$type)) %>% 
  group_split(species)

DV <- readRDS(here("output","dependentVar.rds"))
nrseeds <- length(unique(DV$randomseed))
nrhgtclasses <- length(unique(DV$hgtclass))
```

```{r}
op <- options(pillar.sigfig = 6)
DV %>% 
  filter(randomseed == 1) %>% 
  group_split(species) %>% 
  map2(group_split(IV, species), ~ add_wildlings(.y,.x)) %>% 
  bind_rows() %>% 
  st_drop_geometry() %>% 
  group_by(species) %>% 
  summarize(hectares = sum(are)/100,
            wildlings = sum(wildlings*are)) %>% 
  mutate(per.hectare = wildlings/hectares) # overall wildlings per hectare
options(op)
```

```{r}
IV <- select(IV, -are)

IV <- IV %>% 
  group_split(species) %>% 
  map(~ mutate(., 
               across(c(age,bio01,bio19,relelev), ~ as.vector(scale(.))),
               across(c(seeds.ExP,seeds.WALD), ~ as.vector(scale(log(.), scale = FALSE))))) %>% 
  bind_rows()
length(names(IV)[!names(IV) %in% nottypes]) #number of ecosystems surveyed
```

```{r}
st_union(IV) %>% #to avoid double-counting area that is surveyed for two or more species
  st_area() %>% 
  sum() %>% 
  units::set_units("hectare") #total area surveyed
```

# P.sitchensis-lutzii

```{r}
IVi <- filter(IV, species == "P.sitchensis-lutzii") %>% 
  remove_zero_cols()
summary(IVi)
typenames <- names(IVi)[!names(IVi) %in% nottypes]

DVi <- filter(DV, species == "P.sitchensis-lutzii")
```

## Wildling densities

Height-blind: >= 30cm

```{r}
dat <- add_wildlings(IVi, filter(DVi, randomseed == 1)) %>% 
  st_drop_geometry()

Ps_density <- calculate_wildlings_per_nin(dat, typenames, contrasttype)
```

```{r, echo=FALSE, results='asis'}
Ps_density %>%   
  arrange(desc(density), desc(n), daa) %>% 
  knitr::kable(digits = 1)
```

Height-class 2 >=100cm

```{r}
stratified <- DVi %>% 
  filter(hgtclass == 2) %>%
  select(species, randomseed, hgtclass) %>% 
  nest(dv = geometry) %>% 
  mutate(data = map(dv, ~ add_wildlings(IVi, .))) %>%
  mutate(data = map(data, st_drop_geometry)) %>% 
  select(-dv)

Ps_density_stratified <- stratified %>% 
  mutate(density = map(data, ~ calculate_wildlings_per_nin(., typenames, contrasttype))) %>% 
  select(-data)
```

```{r, include=FALSE}
Ps_density_stratified %>% 
  filter(randomseed == 2) %>%
  select(density) %>% 
  unnest(density) %>% 
  arrange(desc(density), desc(n), daa)
```

## Height-blind model: >= 30cm

Disregards height classes.

Comparing:

1. distribution family (poisson vs. negative binomial vs. generalized poisson)
2. dispersal predictor (seeds.WALD vs. seeds.ExP vs null)

```{r}
newdat <- make_newdat(dat, typenames, contrasttype)
```

### distribution family

```{r}
mf <- paste_modelformula(dat, typenames, contrasttype)
m.poissonWALD <- glmmTMB(mf, dat, offset = seeds.WALD, family = "poisson", 
                         ziformula = ~ age + (1 | locality))
summary(m.poissonWALD)
simulationOutput <- simulateResiduals(fittedModel = m.poissonWALD)
testDispersion(simulationOutput)

m.nbinom2WALD <- glmmTMB(mf, dat, offset = seeds.WALD, family = "nbinom2",
                         ziformula = ~ age + (1 | locality))
summary(m.nbinom2WALD) 
simulationOutput <- simulateResiduals(fittedModel = m.nbinom2WALD)
testDispersion(simulationOutput) 

m.genpoisWALD <- glmmTMB(mf, dat, offset = seeds.WALD, family = "genpois",
                         ziformula = ~ age + (1 | locality))
summary(m.genpoisWALD) 
simulationOutput <- simulateResiduals(fittedModel = m.genpoisWALD)
testDispersion(simulationOutput)
```

### dispersal predictor

```{r}
m.genpoisExP <- glmmTMB(mf, dat, offset = seeds.ExP, family = "genpois",
                        ziformula = ~ age + (1 | locality))
summary(m.genpoisExP)

m.genpoisNULL <- glmmTMB(mf, dat, offset = NULL, family = "genpois",
                         ziformula = ~ age + (1 | locality))
summary(m.genpoisNULL)
```

```{r, include=FALSE}
Ps_dispersalcomp <- bbmle::ICtab(m.genpoisWALD, m.genpoisNULL, base = TRUE, 
                                 delta = TRUE, sort = FALSE) %>% 
  as_tibble() %>% 
  add_column(`seed dispersal estimate` = c('WALD', 'none'), .before = 1) %>% 
  add_row(`seed dispersal estimate` = 'ExP', AIC = NA, dAIC = NA, df = NA, .before = 2) 
```

### Residual diagnostics + predictions

```{r}
simulationOutput <- simulateResiduals(fittedModel = m.genpoisWALD)
plot(simulationOutput)
testOutliers(simulationOutput, type = "bootstrap")
testZeroInflation(simulationOutput)
```

```{r}
Ps_link <- newdat %>% 
  bind_cols(predict(m.genpoisWALD, newdat, se.fit = TRUE, type = "link")) %>% 
  select(locality, fit, se.fit, nin) %>% 
  mutate(species = 'P.sitchensis-lutzii') %>% 
  nest(preds = c(locality, nin, fit, se.fit)) %>% 
  mutate(model = list(m.genpoisWALD))
```

```{r, include=FALSE}
sensmod <- update(m.genpoisWALD, 
                  update.formula(formula(m.genpoisWALD), ~ . + seeds.WALD), 
                  offset = NULL)
newdat <- filter(newdat, is.na(locality))
Ps_cond_sens <- newdat %>% 
  mutate(fit = predict(sensmod, newdat, se.fit = FALSE, type = "conditional")) %>% 
  select(nin, fit)
```

### Types-only null model

```{r}
mf <- formula(paste("wildlings ~", paste(modeltypes, collapse = " + ")))
m.typesonly <- glmmTMB(mf, dat, offset = NULL, family = "genpois",
                       ziformula = ~ age + (1 | locality))
summary(m.typesonly)
```

## Height-class 2 models: >=100 cm

```{r}
stratified <- DVi %>% 
  filter(hgtclass == 2) %>%
  select(species, randomseed, hgtclass) %>% 
  nest(dv = geometry) %>% 
  mutate(data = map(dv, ~ add_wildlings(IVi, .))) %>%
  mutate(data = map(data, st_drop_geometry)) %>% 
  mutate(newdat = map(data, \(x) make_newdat(x, typenames, contrasttype))) %>% 
  mutate(newdat = map(newdat, \(x) filter(x, is.na(locality)))) %>% 
  select(-dv)

possibly_fit <- possibly(fit_genpoisWALDmodel, otherwise = NA)
plan(multisession, workers = availableCores()-1)
stratified <- stratified %>% 
  mutate(model = future_map(data, ~ possibly_fit(., typenames, contrasttype)))

stratified <- stratified %>% 
  filter(!is.na(model)) %>% 
  mutate(preds = future_map2(model, newdat, 
                             \(x, y) predict(x, y, se.fit = TRUE, type = "link"))) %>% 
  mutate(preds = map2(preds, newdat, 
                      \(x, y) bind_cols(x, select(y, locality, nin))))

Ps_stratified <- stratified
```

```{r}
rm(m.poissonWALD, m.nbinom2WALD, m.genpoisWALD, m.genpoisExP, m.genpoisNULL,
   stratified)
```

# P.abies

```{r}
IVi <- filter(IV, species == "P.abies") %>% 
  remove_zero_cols()
summary(IVi)
typenames <- names(IVi)[!names(IVi) %in% nottypes]

DVi <- filter(DV, species == "P.abies")
```

## Wildling densities

Height-blind: >= 30cm

```{r}
dat <- add_wildlings(IVi, filter(DVi, randomseed == 1)) %>% 
  st_drop_geometry()

Pa_density <- calculate_wildlings_per_nin(dat, typenames, contrasttype)
```

```{r, echo=FALSE, results='asis'}
Pa_density %>%   
  arrange(desc(density), desc(n), daa) %>% 
  knitr::kable(digits = 1)
```

Height-class 2 >=100cm

```{r}
stratified <- DVi %>% 
  filter(hgtclass == 2) %>%
  select(species, randomseed, hgtclass) %>% 
  nest(dv = geometry) %>% 
  mutate(data = map(dv, ~ add_wildlings(IVi, .))) %>%
  mutate(data = map(data, st_drop_geometry)) %>% 
  select(-dv)

Pa_density_stratified <- stratified %>% 
  mutate(density = map(data, ~ calculate_wildlings_per_nin(., typenames, contrasttype))) %>%
  select(-data)
```

```{r, include=FALSE}
Pa_density_stratified %>% 
  filter(randomseed == 2) %>%
  select(density) %>% 
  unnest(density) %>% 
  arrange(desc(density), desc(n), daa)
```

## Height-blind model: >= 30cm

Disregards height classes.

Comparing:

1. distribution family (poisson vs. negative binomial vs. generalized poisson)
2. dispersal predictor (seeds.WALD vs. seeds.ExP vs null)

```{r}
newdat <- make_newdat(dat, typenames, contrasttype)
```

### distribution family

```{r}
mf <- paste_modelformula(dat, typenames, contrasttype)
m.poissonWALD <- glmmTMB(mf, dat, offset = seeds.WALD, family = "poisson", 
                         ziformula = ~ age + (1 | locality))
summary(m.poissonWALD)
simulationOutput <- simulateResiduals(fittedModel = m.poissonWALD)
testDispersion(simulationOutput)

m.nbinom2WALD <- glmmTMB(mf, dat, offset = seeds.WALD, family = "nbinom2",
                         ziformula = ~ age + (1 | locality))
summary(m.nbinom2WALD)
simulationOutput <- simulateResiduals(fittedModel = m.nbinom2WALD)
testDispersion(simulationOutput) 

m.genpoisWALD <- glmmTMB(mf, dat, offset = seeds.WALD, family = "genpois",
                         ziformula = ~ age + (1 | locality))
summary(m.genpoisWALD) 
simulationOutput <- simulateResiduals(fittedModel = m.genpoisWALD)
testDispersion(simulationOutput) 
```

### dispersal predictor

```{r}
m.genpoisExP <- glmmTMB(mf, dat, offset = seeds.ExP, family = "genpois",
                        ziformula = ~ age + (1 | locality))
summary(m.genpoisExP)

m.genpoisNULL <- glmmTMB(mf, dat, offset = NULL, family = "genpois",
                         ziformula = ~ age + (1 | locality))
summary(m.genpoisNULL)
```

```{r, include=FALSE}
Pa_dispersalcomp <- bbmle::ICtab(m.genpoisWALD, m.genpoisNULL, base = TRUE, 
                                 delta = TRUE, sort = FALSE) %>% 
  as_tibble() %>% 
  add_column(`seed dispersal estimate` = c('WALD', 'none'), .before = 1) %>% 
  add_row(`seed dispersal estimate` = 'ExP', AIC = NA, dAIC = NA, df = NA, .before = 2) 
```

### Residual diagnostics + predictions

```{r}
simulationOutput <- simulateResiduals(fittedModel = m.genpoisWALD)
plot(simulationOutput)
testOutliers(simulationOutput, type = "bootstrap")
testZeroInflation(simulationOutput)
```

```{r}
Pa_link <- newdat %>% 
  bind_cols(predict(m.genpoisWALD, newdat, se.fit = TRUE, type = "link")) %>% 
  select(locality, fit, se.fit, nin) %>% 
  mutate(species = 'P.abies') %>% 
  nest(preds = c(locality, nin, fit, se.fit)) %>% 
  mutate(model = list(m.genpoisWALD))
```

```{r, include=FALSE}
sensmod <- update(m.genpoisWALD, 
                  update.formula(formula(m.genpoisWALD), ~ . + seeds.WALD), 
                  offset = NULL)
newdat <- filter(newdat, is.na(locality))
Pa_cond_sens <- newdat %>% 
  mutate(fit = predict(sensmod, newdat, se.fit = FALSE, type = "conditional")) %>% 
  select(nin, fit)
```

### Types-only null model

```{r}
mf <- formula(paste("wildlings ~", paste(modeltypes, collapse = " + ")))
m.typesonly <- glmmTMB(mf, dat, offset = NULL, family = "genpois",
                       ziformula = ~ age + (1 | locality))
summary(m.typesonly)
```

## Height-class 2 models: >=100 cm

```{r}
stratified <- DVi %>% 
  filter(hgtclass == 2) %>%
  select(species, randomseed, hgtclass) %>% 
  nest(dv = geometry) %>% 
  mutate(data = map(dv, ~ add_wildlings(IVi, .))) %>%
  mutate(data = map(data, st_drop_geometry)) %>% 
  mutate(newdat = map(data, \(x) make_newdat(x, typenames, contrasttype))) %>% 
  mutate(newdat = map(newdat, \(x) filter(x, is.na(locality)))) %>% 
  select(-dv)

possibly_fit <- possibly(fit_genpoisWALDmodel, otherwise = NA)
plan(multisession, workers = availableCores()-1)
stratified <- stratified %>% 
  mutate(model = future_map(data, ~ possibly_fit(., typenames, contrasttype)))

stratified <- stratified %>% 
  filter(!is.na(model)) %>% 
  mutate(preds = future_map2(model, newdat, 
                             \(x, y) predict(x, y, se.fit = TRUE, type = "link"))) %>% 
  mutate(preds = map2(preds, newdat, 
                      \(x, y) bind_cols(x, select(y, locality, nin))))

Pa_stratified <- stratified
```

```{r}
rm(m.poissonWALD, m.nbinom2WALD, m.genpoisWALD, m.genpoisExP, m.genpoisNULL,
   stratified)
```

# Larix

```{r}
IVi <- filter(IV, species == "Larix") %>% 
  remove_zero_cols()
summary(IVi)
typenames <- names(IVi)[!names(IVi) %in% nottypes]

DVi <- filter(DV, species == "Larix")
```

## Wildling densities

Height-blind: >= 30cm

```{r}
dat <- add_wildlings(IVi, filter(DVi, randomseed == 1)) %>% 
  st_drop_geometry()

L_density <- calculate_wildlings_per_nin(dat, typenames, contrasttype)
```

```{r, echo=FALSE, results='asis'}
L_density %>%   
  arrange(desc(density), desc(n), daa) %>% 
  knitr::kable(digits = 1)
```

Height-class 2 >=100cm

```{r}
stratified <- DVi %>% 
  filter(hgtclass == 2) %>%
  select(species, randomseed, hgtclass) %>% 
  nest(dv = geometry) %>% 
  mutate(data = map(dv, ~ add_wildlings(IVi, .))) %>%
  mutate(data = map(data, st_drop_geometry)) %>% 
  select(-dv)

L_density_stratified <- stratified %>% 
  mutate(density = map(data, ~ calculate_wildlings_per_nin(., typenames, contrasttype))) %>%
  select(-data)
```

```{r, include=FALSE}
L_density_stratified %>% 
  filter(randomseed == 2) %>%
  select(density) %>% 
  unnest(density) %>% 
  arrange(desc(density), desc(n), daa)
```

## Height-blind model: >= 30cm

Disregards height classes.

Comparing:

1. distribution family (poisson vs. negative binomial vs. generalized poisson)
2. dispersal predictor (seeds.WALD vs. seeds.ExP vs null)

```{r}
newdat <- make_newdat(dat, typenames, contrasttype)
```

### distribution family

```{r}
mf <- paste_modelformula(dat, typenames, contrasttype)
m.poissonWALD <- glmmTMB(mf, dat, offset = seeds.WALD, family = "poisson", 
                         ziformula = ~ age + (1 | locality))
summary(m.poissonWALD)
simulationOutput <- simulateResiduals(fittedModel = m.poissonWALD)
testDispersion(simulationOutput)

m.nbinom2WALD <- glmmTMB(mf, dat, offset = seeds.WALD, family = "nbinom2",
                         ziformula = ~ age + (1 | locality))
summary(m.nbinom2WALD)
simulationOutput <- simulateResiduals(fittedModel = m.nbinom2WALD)
testDispersion(simulationOutput)

m.genpoisWALD <- glmmTMB(mf, dat, offset = seeds.WALD, family = "genpois",
                         ziformula = ~ age + (1 | locality))
summary(m.genpoisWALD)
simulationOutput <- simulateResiduals(fittedModel = m.genpoisWALD)
testDispersion(simulationOutput) 
```

### dispersal predictor

```{r}
m.genpoisExP <- glmmTMB(mf, dat, offset = seeds.ExP, family = "genpois",
                        ziformula = ~ age + (1 | locality))
summary(m.genpoisExP)

m.genpoisNULL <- glmmTMB(mf, dat, offset = NULL, family = "genpois",
                        ziformula = ~ age + (1 | locality))
summary(m.genpoisNULL)
```

```{r, include=FALSE}
L_dispersalcomp <- bbmle::ICtab(m.genpoisWALD, m.genpoisExP, base = TRUE, 
                                delta = TRUE, sort = FALSE) %>% 
  as_tibble() %>% 
  add_column(`seed dispersal estimate` = c('WALD', 'ExP'), .before = 1) %>% 
  add_row(`seed dispersal estimate` = 'none', AIC = NA, dAIC = NA, df = NA) 
```

### Residual diagnostics + predictions

```{r}
simulationOutput <- simulateResiduals(fittedModel = m.genpoisWALD)
plot(simulationOutput)
testOutliers(simulationOutput, type = "bootstrap")
testZeroInflation(simulationOutput)
```

```{r}
L_link <- newdat %>% 
  bind_cols(predict(m.genpoisWALD, newdat, se.fit = TRUE, type = "link")) %>% 
  select(locality, fit, se.fit, nin) %>% 
  mutate(species = 'Larix') %>% 
  nest(preds = c(locality, nin, fit, se.fit)) %>% 
  mutate(model = list(m.genpoisWALD))
```

```{r, include=FALSE}
sensmod <- update(m.genpoisWALD, 
                  update.formula(formula(m.genpoisWALD), ~ . + seeds.WALD), 
                  offset = NULL)
newdat <- filter(newdat, is.na(locality))
L_cond_sens <- newdat %>% 
  mutate(fit = predict(sensmod, newdat, se.fit = FALSE, type = "conditional")) %>% 
  select(nin, fit)
```

### Types-only null model

```{r}
mf <- formula(paste("wildlings ~", paste(modeltypes, collapse = " + ")))
m.typesonly <- glmmTMB(mf, dat, offset = NULL, family = "genpois",
                       ziformula = ~ age + (1 | locality))
summary(m.typesonly)
```

## Height-class 2 models: >=100 cm

```{r}
stratified <- DVi %>% 
  filter(hgtclass == 2) %>%
  select(species, randomseed, hgtclass) %>% 
  nest(dv = geometry) %>% 
  mutate(data = map(dv, ~ add_wildlings(IVi, .))) %>%
  mutate(data = map(data, st_drop_geometry)) %>% 
  mutate(newdat = map(data, \(x) make_newdat(x, typenames, contrasttype))) %>% 
  mutate(newdat = map(newdat, \(x) filter(x, is.na(locality)))) %>% 
  select(-dv)

possibly_fit <- possibly(fit_genpoisWALDmodel, otherwise = NA)
plan(multisession, workers = availableCores()-1)
stratified <- stratified %>% 
  mutate(model = future_map(data, ~ possibly_fit(., typenames, contrasttype)))

stratified <- stratified %>% 
  filter(!is.na(model)) %>% 
  mutate(preds = future_map2(model, newdat, 
                             \(x, y) predict(x, y, se.fit = TRUE, type = "link"))) %>% 
  mutate(preds = map2(preds, newdat, 
                      \(x, y) bind_cols(x, select(y, locality, nin))))

L_stratified <- stratified
```

```{r}
rm(m.poissonWALD, m.nbinom2WALD, m.genpoisWALD, m.genpoisExP, m.genpoisNULL,
   stratified)
```

# All species

```{r}
species <- c("P.sitchensis-lutzii", "P.abies", "Larix")

list(Ps_density, Pa_density, L_density) %>% 
  set_names(species) %>% 
  bind_rows(.id = "species") %>% 
  nest(density = c(nin, daa, n, density, fit)) %>% 
  bind_rows(Ps_density_stratified, Pa_density_stratified, L_density_stratified) %>%
  saveRDS(here("output","results-densities.rds"))

list(Ps_dispersalcomp, Pa_dispersalcomp, L_dispersalcomp) %>% 
  set_names(species) %>% 
  saveRDS(here("output","results-dispersalcomparison.rds"))

bind_rows(Ps_link, Pa_link, L_link) %>% 
  bind_rows(Ps_stratified, Pa_stratified, L_stratified) %>%
  mutate(preds = map(preds, function(x) {
    x <- mutate(x, ci.l = fit - 1.96*se.fit, ci.u = fit + 1.96*se.fit) %>% 
      mutate(across(c('fit', 'ci.l', 'ci.u'), ~ exp(.))) %>% 
      select(-se.fit) 
    scaler <- filter(x, nin == contrasttype, is.na(locality)) %>% 
      pull(fit)
    mutate(x, across(c('fit', 'ci.l', 'ci.u'), ~ `/`(., scaler)))
  })) %>%
  saveRDS(here("output","results-models.rds"))
```

```{r, include=FALSE}
pred_sens.tbl <- list(Ps_cond_sens, Pa_cond_sens, L_cond_sens) %>% 
  set_names(species) %>% 
  bind_rows(.id = "species") %>% 
  group_by(species) %>% 
  group_split(.keep = TRUE) %>% 
  map(function(x) {
    scaler <- filter(x, nin == contrasttype) %>% 
      pull(fit)
    mutate(x, fit = fit/scaler)
  }) %>% 
  bind_rows()

saveRDS(pred_sens.tbl, here("output","results-sensitivity-offsetting.rds"))
```

```{r sessionInfo}
sessionInfo()
```
