---
title: "Modeling"
output: 
  html_document:
    toc: true
    toc_float: true
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(here)
library(tidyverse)
library(glmmTMB)
library(DHARMa)
library(emmeans)
```

```{r}
datlist <- readRDS(here("output","data.rds"))
```

# Larix

```{r}
L_dat <- Filter(function(x)!all(is.na(x)), datlist$Larix)

excluded <- "plantation forest"
perfectlyseparated <- c('T18', 'T30', 'T2', 'T27', 'swamp', 'T13')
removed <- c(excluded, perfectlyseparated)
L_dat <- L_dat %>% 
  filter_at(removed, all_vars(. < 0.5)) %>% 
  filter(!(is.na(seeds.ExP) | is.na(seeds.ExP))) %>% 
  select(-c('species')) %>% 
  psycho::standardize(subset = list("seeds.WALD","seeds.ExP","age",
                                    "bio01","bio19","relelev"))
summary(L_dat)

contrast <- "T4"
paste(names(L_dat)[!(names(L_dat) %in% c(contrast, removed))], collapse = " + ")
```

Perfectly separated: T18, T30, T2, T27, swamp, T13 (0 wildlings, from wildlings-NiN.Rmd)

```{r}
L_arith <- pivot_longer(L_dat, 9:ncol(L_dat), names_to = "nin", values_to = "are") %>%
  group_by(nin) %>% 
  summarize(daa = sum(are)/10, # units from are to decare
            n = sum(wildlings*are)) %>% 
  mutate(density = n/daa) %>%
  arrange(desc(density), desc(n), daa)
scaler <- filter(L_arith, nin == contrast) %>% 
  pull(density)
L_arith <- mutate(L_arith, fit = density/scaler)
```

### Compare WALD vs ExP

```{r}
L_f1 <- formula("wildlings ~ seeds.WALD * age + bio01 + bio19 + relelev + cultivated + disturbed + T32 + fen + T34 + T1 + T17 + (1 | locality)")
L_m1 <- glmmTMB(L_f1, L_dat, family = "nbinom2", ziformula = ~ 0)
summary(L_m1)
```

```{r}
L_f2 <- formula("wildlings ~ seeds.ExP * age + bio01 + bio19 + relelev + cultivated + disturbed + T32 + fen + T34 + T1 + T17 + (1 | locality)")
L_m2 <- glmmTMB(L_f2, L_dat, family = "nbinom2", ziformula = ~ 0)
summary(L_m2)
```

```{r}
bbmle::AICtab(L_m1, L_m2)
```

Continue with WALD.

### Compare zero inflation effects

```{r}
L_m3 <- glmmTMB(L_f1, L_dat, family = "nbinom2", ziformula = ~ 1)
summary(L_m3)
```

```{r}
L_m4 <- glmmTMB(L_f1, L_dat, family = "nbinom2", ziformula = ~ age)
summary(L_m4)
```

```{r}
L_m5 <- glmmTMB(L_f1, L_dat, family = "nbinom2", ziformula = ~ age + (1 | locality))
summary(L_m5)
```

```{r}
bbmle::AICtab(L_m1, L_m3, L_m4, L_m5)
```

m5 is better than m1, m3, and m4: much better AIC, clear zero-inflation effects

```{r}
simulationOutput <- simulateResiduals(fittedModel = L_m5)
plot(simulationOutput)
testDispersion(simulationOutput)
testZeroInflation(simulationOutput)
```

Diagnostics look OK.

### Interpret relative establishment probability

```{r}
L_m5$frame %>% map_dbl(mean)

nd <- expand_grid(seeds.WALD = 0, age = 0, bio01 = 0, bio19 = 0, relelev = 0,
                  cultivated = c(0,1), disturbed = c(0,1), T32 = c(0,1), 
                  fen = c(0,1), T34 = c(0,1), T1 = c(0,1), T17 = c(0,1), 
                  locality = NA_real_) %>% 
  mutate(sum = cultivated + disturbed + T32 + fen + T34 + T1 + T17) %>% 
  filter(sum %in% c(0,1)) %>% 
  select(-sum)

link <- bind_cols(nd, predict(L_m5, nd, se.fit = TRUE, type = "link")); link
resp <- bind_cols(nd, predict(L_m5, nd, se.fit = TRUE, type = "response")); resp
cond <- bind_cols(nd, predict(L_m5, nd, se.fit = TRUE, type = "conditional")); cond
cond <- mutate(cond, se.l = fit - se.fit, se.u = fit + se.fit) %>% 
  mutate_at(.vars = c('fit', 'se.l', 'se.u'), ~ log(.)) %>% 
  mutate(ci.l = fit + 1.96*(se.l - fit),
         ci.u = fit + 1.96*(se.u - fit)) %>% 
  mutate_at(.vars = c('fit', 'se.l', 'se.u', 'ci.l', 'ci.u'), ~ exp(.))
  
# ndz1 <- expand_grid(seeds.WALD = -1, age = -1, bio01 = -1, bio19 = -1, relelev = -1,
#                       cultivated = c(0,1), disturbed = c(0,1), T32 = c(0,1), 
#                       fen = c(0,1), T34 = c(0,1), T1 = c(0,1), T17 = c(0,1), 
#                       locality = NA_real_) %>% 
#   mutate(sum = cultivated + disturbed + T32 + fen + T34 + T1 + T17) %>% 
#   filter(sum %in% c(0,1)) %>% 
#   select(-sum) 
# condz1 <- cbind(ndz1, predict(L_m5, ndz1, se.fit = TRUE, type = "conditional")); condz1

cond <- mutate(cond, nin = names(cond[6:12])[apply(cond[6:12], 1, match, x = 1)]) %>% 
  replace_na(list(nin = "T4"))
scaler <- filter(cond, nin == "T4") %>% 
  pull(fit)
cond.scaled <- mutate_at(cond, 14:19, ~  `/`(., scaler))

```

Note: ratios of resp estimates are identical to ratios of cond estimates

```{r}
gg <- ggplot(cond.scaled, aes(nin, fit)) + 
  geom_linerange(mapping = aes(ymin = ci.l, ymax = ci.u)) + 
  geom_point()  + 
  geom_point(data = filter(L_arith, nin %in% cond.scaled$nin), 
             col = "grey", pch = 1)
gg + scale_y_log10() + coord_flip()
```


# P.abies

```{r}
Pa_dat <- Filter(function(x)!all(is.na(x)), datlist$P.abies)

excluded <- "plantation forest"
perfectlyseparated <- c('T6', 'spring', 'T31', 'T2')
removed <- c(excluded, perfectlyseparated)
Pa_dat <- Pa_dat %>% 
  filter_at(removed, all_vars(. < 0.5)) %>% 
  filter(!(is.na(seeds.ExP) | is.na(seeds.ExP))) %>% 
  select(-c('species')) %>% 
  psycho::standardize(subset = list("seeds.WALD","seeds.ExP","age",
                                    "bio01","bio19","relelev"))
summary(Pa_dat)

contrast <- "T4"
paste(names(Pa_dat)[!(names(Pa_dat) %in% c(contrast, removed))], collapse = " + ")
```

Perfectly separated: T6, spring, T31, T2

```{r}
Pa_arith <- pivot_longer(Pa_dat, 9:ncol(Pa_dat), names_to = "nin", values_to = "are") %>%
  group_by(nin) %>% 
  summarize(daa = sum(are)/10, # units from are to decare
            n = sum(wildlings*are)) %>% 
  mutate(density = n/daa) %>%
  arrange(desc(density), desc(n), daa)
scaler <- filter(Pa_arith, nin == contrast) %>% 
  pull(density)
Pa_arith <- mutate(Pa_arith, fit = density/scaler)
```

### Compare WALD vs ExP

```{r}
Pa_f1 <- formula("wildlings ~ seeds.WALD * age + bio01 + bio19 + relelev + cultivated + disturbed + T32 + fen + T1 + swamp + bog + (1 | locality)")
Pa_m1 <- glmmTMB(Pa_f1, Pa_dat, family = "nbinom2", ziformula = ~ 0)
summary(Pa_m1)
```

```{r}
Pa_f2 <- formula("wildlings ~ seeds.ExP * age + bio01 + bio19 + relelev + cultivated + disturbed + T32 + fen + T1 + swamp + bog + (1 | locality)")
Pa_m2 <- glmmTMB(Pa_f2, Pa_dat, family = "nbinom2", ziformula = ~ 0)
summary(Pa_m2)
```

```{r}
bbmle::AICtab(Pa_m1, Pa_m2)
```

Continue with WALD.

### Compare zero inflation effects

```{r}
Pa_m3 <- glmmTMB(Pa_f1, Pa_dat, family = "nbinom2", ziformula = ~ 1)
summary(Pa_m3)
```

```{r}
Pa_m4 <- glmmTMB(Pa_f1, Pa_dat, family = "nbinom2", ziformula = ~ age)
summary(Pa_m4)
```

```{r}
Pa_m5 <- glmmTMB(Pa_f1, Pa_dat, family = "nbinom2", ziformula = ~ age + (1 | locality))
summary(Pa_m5)
```

```{r}
bbmle::AICtab(Pa_m1, Pa_m3, Pa_m4, Pa_m5)
```

m5 is better than m1, m3, and m4: much better AIC, zero-inflation effects

Drop insignificant interaction?

```{r}
Pa_f6 <- update.formula(Pa_f1, ~ . - seeds.WALD:age)
Pa_m6 <- glmmTMB(Pa_f6, Pa_dat, family = "nbinom2", ziformula = ~ age + (1 | locality))
summary(Pa_m6)
```

```{r}
bbmle::AICtab(Pa_m5, Pa_m6)
```

```{r}
simulationOutput <- simulateResiduals(fittedModel = Pa_m6)
plot(simulationOutput)
testDispersion(simulationOutput)
testZeroInflation(simulationOutput)
```

Diagnostics look bad. Dispersion test p-value = 0.02. Zero-inflation test p-value = 0.38.

```{r}
Pa_m7 <- glmmTMB(Pa_f6, Pa_dat, family = "genpois", ziformula = ~ age + (1 | locality))
summary(Pa_m7)
```

```{r}
simulationOutput <- simulateResiduals(fittedModel = Pa_m7)
plot(simulationOutput)
testDispersion(simulationOutput)
testZeroInflation(simulationOutput)
```

Better diagnostics with generalized poisson. Dispersion test p-value = 0.12. Zero-inflation test p-value = 0.55.

```{r}
Pa_m8 <- glmmTMB(Pa_f6, Pa_dat, family = poisson, ziformula = ~ age + (1 | locality))
summary(Pa_m8)
```

```{r}
simulationOutput <- simulateResiduals(fittedModel = Pa_m8)
plot(simulationOutput)
testDispersion(simulationOutput)
testZeroInflation(simulationOutput)
```

Better diagnostics with poisson. Dispersion test p-value = 0.34. Zero-inflation test p-value = 0.64.

```{r}
bbmle::AICtab(Pa_m6, Pa_m7, Pa_m8)
```

Continue with generalized poisson model: best AIC and OK diagnostics. 

### Interpret relative establishment probability

```{r}
Pa_m7$frame %>% map_dbl(mean)

nd <- expand_grid(seeds.WALD = 0, age = 0, bio01 = 0, bio19 = 0, relelev = 0,
                  cultivated = c(0,1), disturbed = c(0,1), T32 = c(0,1), 
                  fen = c(0,1), T1 = c(0,1), swamp = c(0,1), bog = c(0,1), 
                  locality = NA_real_) %>% 
  mutate(sum = cultivated + disturbed + T32 + fen + T1 + swamp + bog) %>% 
  filter(sum %in% c(0,1)) %>% 
  select(-sum)

link <- bind_cols(nd, predict(Pa_m7, nd, se.fit = TRUE, type = "link")); link
link <- mutate(link, ci.l = fit - 1.96*se.fit, ci.u = fit + 1.96*se.fit) %>% 
  select(-se.fit) %>% 
  mutate_at(c('fit', 'ci.l', 'ci.u'), ~ exp(.))
link <- mutate(link, nin = names(link[6:12])[apply(link[6:12], 1, match, x = 1)]) %>% 
  replace_na(list(nin = "T4"))
scaler <- filter(link, nin == "T4") %>% 
  pull(fit)
link.scaled <- mutate_at(link, c('fit', 'ci.l', 'ci.u'), ~  `/`(., scaler))

cond <- bind_cols(nd, predict(Pa_m7, nd, se.fit = TRUE, type = "conditional")); cond
cond <- mutate(cond, se.l = fit - se.fit, se.u = fit + se.fit) %>% 
  mutate_at(.vars = c('fit', 'se.l', 'se.u'), ~ log(.)) %>% 
  mutate(ci.l = fit + 1.96*(se.l - fit),
         ci.u = fit + 1.96*(se.u - fit)) %>% 
  mutate_at(.vars = c('fit', 'se.l', 'se.u', 'ci.l', 'ci.u'), ~ exp(.))
cond <- mutate(cond, nin = names(cond[6:12])[apply(cond[6:12], 1, match, x = 1)]) %>% 
  replace_na(list(nin = "T4"))
scaler <- filter(cond, nin == "T4") %>% 
  pull(fit)
cond.scaled <- mutate_at(cond, 14:19, ~  `/`(., scaler))

```

Note: ratios of resp estimates are identical to ratios of cond estimates

```{r}
gg <- ggplot(cond.scaled, aes(nin, fit)) + 
  geom_linerange(mapping = aes(ymin = ci.l, ymax = ci.u)) + 
  geom_point()  + 
  geom_point(data = filter(Pa_arith, nin %in% cond.scaled$nin), 
             col = "grey", pch = 1)
gg + scale_y_log10() + coord_flip()
```

```{r}
gg <- ggplot(link.scaled, aes(nin, fit)) + 
  geom_linerange(mapping = aes(ymin = ci.l, ymax = ci.u)) + 
  geom_point()  + 
  geom_point(data = filter(Pa_arith, nin %in% cond.scaled$nin), 
             col = "grey", pch = 1)
gg + scale_y_log10() + coord_flip()
```



# P.abies

```{r}
Pa_dat <- datlist$P.abies %>% 
  psycho::standardize(subset = list("seeds.WALD","seeds.ExP","age",
                                    "bio01","bio19","relelev"))
str(Pa_dat)
paste(names(Pa_dat), collapse = " + ")
```

Perfectly separated: T6, spring, T31, T2

```{r}
Pa_f1 <- formula("wildlings ~ seeds.WALD + age + bio01 + bio19 + relelev + cultivated + disturbed + T32 + fen + T4 + T1 + swamp + bog + (1 | locality)")
Pa_m1 <- glmmTMB(Pa_f1, Pa_dat, family = "nbinom2", ziformula = ~1, verbose = TRUE)
summary(Pa_m1)
```

```{r}
Pa_fzi2 <- formula("~ age")
Pa_m2 <- glmmTMB(Pa_f1, Pa_dat, family = "nbinom2", ziformula = Pa_fzi2, verbose = TRUE)
summary(Pa_m2)
ranef(Pa_m2)

bbmle::AICtab(Pa_m1, Pa_m2)
```

```{r}
simulationOutput <- simulateResiduals(fittedModel = Pa_m2)
plot(simulationOutput) # Highest predicted values (from ~90th percentile)
testDispersion(simulationOutput)
testZeroInflation(simulationOutput)
```

The model's highest predicted values (from ~90th percentile and up) tend to underestimate the observed values --- but not to a degree that look egregious. 

Could this be due to spatial autocorrelation at a scale smaller than the minimum polygon size?

# P.contorta

```{r}
Pc_dat <- datlist$P.contorta %>% 
  psycho::standardize(subset = list("seeds.WALD","seeds.ExP","age",
                                    "bio01","bio19","relelev"))
str(Pc_dat)
paste(names(Pc_dat)[map_lgl(Pc_dat, ~ !all(is.na(.)))], collapse = " + ")
```

Perfectly separated: cultivated, T27, T32

```{r}
Pc_f1 <- formula("wildlings ~ seeds.WALD + age + bio01 + bio19 + relelev + disturbed + fen + T34 + T4 + T1 + swamp + T30 + bog + (1 | locality)")
Pc_m1 <- glmmTMB(Pc_f1, Pc_dat, family = "nbinom2", ziformula = ~1, verbose = TRUE)
summary(Pc_m1)
```


# P.sitchensis


```{r sessionInfo}
sessionInfo()
```