---
title: "Manuscript figures"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(patchwork)
library(ggtext)
library(here)
library(sf)
library(rnaturalearth)
source(here("R","functions.R"))
```

# Sites map

```{r, eval=FALSE}
ne_download(scale = 50, type = 'populated_places', category = 'cultural', 
            destdir = here("data", "raw", "cartographic"), load = FALSE)
ne_download(scale = 10, type = 'minor_islands', category = 'physical', 
            destdir = here("data", "raw", "cartographic"), load = FALSE)
```

```{r}
cities <- ne_load(scale = 50, type = 'populated_places', category = 'cultural', 
                  destdir = here("data", "raw", "cartographic"), returnclass = 'sf')
norcities <- filter(cities, NAME %in% c("Oslo", "Trondheim", "Bergen"))
norislands <- ne_load(scale = 10, type = 'minor_islands', category = 'physical', 
                destdir = here("data", "raw", "cartographic"), returnclass = 'sf') %>% 
  st_crop(xmin = 4, xmax = 32, ymin = 58, ymax = 72) %>% 
  st_transform(crs = 25833) # ETRS89 / UTM zone 33N

nor50 <- ne_countries(scale = 50, country = c("Norway"), returnclass = "sf") %>% 
  st_transform(crs = 25833)
nor10 <- ne_countries(scale = 10, country = "Norway", returnclass = "sf") %>% 
  st_transform(crs = 25833)

locs <- read_csv(here("data", "localities.csv"), col_types = cols()) %>% 
  st_as_sf(coords=c("utm33.easting", "utm33.northing"), crs = 25833)
locs <- mutate(locs, species = factor(species, levels = c("P.sitchensis-lutzii",
                                                          "P.abies",
                                                          "Larix")))
st_bbox(locs)

filter(locs, county %in% c("Rogaland","Hordaland")) %>% 
  st_bbox()
xlimits <- c(-69e3,10e3)
ylimits <- c(6.515e6,6.775e6)
inset <- st_polygon(list(matrix(c(xlimits[1],ylimits[1],
                         xlimits[2],ylimits[1],
                         xlimits[2],ylimits[2],
                         xlimits[1],ylimits[2],
                         xlimits[1],ylimits[1]), byrow = T, ncol = 2))) %>% 
  st_sfc(crs = 25833)

filter(locs, year.registered == 2019) %>% 
  st_bbox()
xlimits2 <- c(362e3,430e3)
ylimits2 <- c(7.255e6,7.355e6)
inset2 <- st_polygon(list(matrix(c(xlimits2[1],ylimits2[1],
                         xlimits2[2],ylimits2[1],
                         xlimits2[2],ylimits2[2],
                         xlimits2[1],ylimits2[2],
                         xlimits2[1],ylimits2[1]), byrow = T, ncol = 2))) %>% 
  st_sfc(crs = 25833)
CBdark2.3class <- c('#d95f02', '#7570b3', '#1b9e77')

g1 <- ggplot() +
  geom_sf(data = nor50, fill = "white") + 
  geom_sf(data = norcities) +
  geom_sf_text(data = norcities, aes(label = NAME), 
               nudge_y = c(-3e4,3e4,4e4), 
               nudge_x = c(0, 10e4, 0), size = 2) +
  geom_sf(data = locs, #st_jitter(locs, amount=5e3)
          mapping = aes(color = species), 
          show.legend = "point", size = 1.5, alpha = 0.6) +
  geom_sf(data = inset, fill = NA, linetype = "longdash", color = 'black') + 
  geom_sf(data = inset2, fill = NA, linetype = "longdash", color = 'black') + 
  scale_color_manual(labels = c("<span style = 'color:#d95f02;'>*Picea sitchensis* / 
       *Picea* \u00D7 *lutzii*</span>", 
       "<span style = 'color:#7570b3;'>*Picea abies*</span>", 
       "<span style = 'color:#1b9e77;'>*Larix* spp.</span>"), values = CBdark2.3class) +
  coord_sf(xlim = c(-60e3,730e3), ylim = c(6.5e6,7.8e6)) +
  theme_minimal() +
  theme(legend.position = c(0.42,0.9), legend.title = element_blank(), 
        legend.text = element_markdown(),
        axis.title.x=element_blank(), axis.title.y=element_blank(),
        panel.background = element_rect(fill = "ghostwhite"))

g2 <- ggplot() +
  geom_sf(data = nor10, fill = "white") + 
  geom_sf(data = norislands, fill = "white") + 
  geom_sf(data = norcities) +
  geom_sf(data = locs, 
          mapping = aes(color = species.latin), 
          show.legend = FALSE, size = 2, alpha = 0.7) + 
  scale_color_manual(values = CBdark2.3class) + 
  coord_sf(xlim = xlimits, ylim = ylimits, datum = NA) +
  theme_minimal() +
  theme(plot.margin = margin(0, 0, 0, 0),
        panel.border = element_rect(fill = NA, size=1),
        panel.background = element_rect(fill = "ghostwhite"))

g3 <- ggplot() +
  geom_sf(data = nor10, fill = "white") + 
  geom_sf(data = norislands, fill = "white") + 
  geom_sf(data = locs, 
          mapping = aes(color = species.latin), 
          show.legend = FALSE, size = 2, alpha = 0.7) + 
  scale_color_manual(values = CBdark2.3class) +
  coord_sf(xlim = xlimits2, ylim = ylimits2, datum = NA) +
  theme_minimal() +
  theme(plot.margin = margin(0, 0, 0, 0),
        panel.border = element_rect(color = "black", fill = NA, size=1),
        panel.background = element_rect(fill = "ghostwhite"))

scalefactor <- 3.6
g2x <- diff(xlimits)
g2y <- diff(ylimits)
ggplot_build(g1)$layout$coord$limits
xmax <- 750e3
ymin <- 6.45e6
g12 <- g1 + 
  annotation_custom(ggplotGrob(g2), 
                    xmin = xmax - scalefactor*g2x, xmax = xmax, 
                    ymin = ymin, ymax = ymin + scalefactor*g2y) +
  annotate("segment", linetype = "dotted",
           x = xlimits[2], xend = xmax - scalefactor*g2x, 
           y = ylimits[1], yend = ymin)
  
g2x <- diff(xlimits2)
g2y <- diff(ylimits2)
st_bbox(locs)
xmin <- -80e3
ymin <- 7.11e6
g123 <- g12 + 
  annotation_custom(ggplotGrob(g3), 
                                xmin = xmin, xmax = xmin + scalefactor*g2x, 
                                ymin = ymin, ymax = ymin + scalefactor*g2y) +
  annotate("segment", linetype = "dotted",
           x = xlimits2[1], xend = xmin + scalefactor*g2x, 
           y = ylimits2[2], yend = ymin + scalefactor*g2y)
```

```{r}
ggsave(g123, filename = 'sites-map.svg', path = here('ms', 'figures'),
  scale = 1, width = 74, height = 105, units = 'mm', dpi = 300)
```

# Site example

Criteria for selecting example location:

- Sitka spruce
- Has mosaic types
- Sources mapped outside 500x500 m

Of those locations that fill these criteria, chose that with most wildlings: Meaasen (2019).

```{r}
types <- read_csv(here('data','types.csv'))
IV <- readRDS(here("output","independentVars.rds")) %>% 
  st_as_sf() %>% 
  filter(species == "P.sitchensis-lutzii", locality == "meaasen") %>% 
  filter(`plantation forest` < 0.5) %>% 
  rowwise() %>% 
  mutate(are = sum(c_across(any_of(types$type)))) %>% 
  filter(are >= 0.5) %>% 
  ungroup()
DV <- readRDS(here("output","dependentVar.rds")) %>% 
  filter(species == "P.sitchensis-lutzii", locality == "meaasen", 
         randomseed == 2, hgtclass == 2)
dat <- add_wildlings(IV, DV)
st_bbox(dat)
```

```{r}
png(filename = here('ms', 'figures', 'site-example', 'wildlings.png'),
    width = 10, height = 10, units = 'cm', res = 300)
plot(dat["wildlings"], main = 'wildlings, height \u2265 100 cm', 
     key.pos = 1, key.width = lcm(1.4))
dev.off()

png(filename = here('ms', 'figures', 'site-example', 'WALD.png'),
    width = 10, height = 10, units = 'cm', res = 300)
plot(dat["seeds.WALD"], main = 'WALD seed probability density', 
     key.pos = 1, key.width = lcm(1.4), logz = TRUE, at = c(-5,-4,-3))
dev.off()

png(filename = here('ms', 'figures', 'site-example', 'T2.png'),
    width = 10, height = 10, units = 'cm', res = 300)
plot(dat["T2"], main = 'open shallow-soil ground', 
     key.pos = 1, key.width = lcm(1.4), nbreaks = 6)
dev.off()

png(filename = here('ms', 'figures', 'site-example', 'T4.png'),
    width = 10, height = 10, units = 'cm', res = 300)
plot(dat["T4"], main = 'forest', 
     key.pos = 1, key.width = lcm(1.4), nbreaks = 6)
dev.off()
```

# Wildling abundance by ecosystem type

```{r}
results.models <- readRDS(here("output","results-models.rds"))
results.densities <- readRDS(here("output","results-densities.rds"))
results.sensitivity.offsetting <- readRDS(here("output","results-sensitivity-offsetting.rds"))
```

```{r}
types <- read_csv(here('data', 'types.csv')) %>% 
  mutate(name = ifelse(category == 'wetland', paste0(paste0('*',name),'*'), name)) %>%
  mutate(name.br = case_when(
    name == 'drained wetland and terrestrialized freshwater sediment' ~
      'drained wetland and<br>terrestrialized freshwater sediment',
    name == 'strongly altered ground with semi-natural grassland character' ~
      'strongly altered ground with<br>semi-natural grassland character',
    name == 'agriculturally improved grassland with semi-natural character' ~
      'agriculturally improved grassland<br>with semi-natural character',
    TRUE ~ name
  )) %>% 
  rename(nin = type) %>% 
  select(-vegheight, -category, -structuring)

topbottomtheme <- theme_light() + 
  theme(plot.title = element_markdown(),
        axis.title.x = element_markdown(),
        axis.text.y = element_markdown(),
        axis.title.y = element_blank(),
        axis.ticks = element_blank(),
        strip.background =element_rect(fill="lightgrey"),
        strip.text = element_text(colour = "black"),
        panel.spacing = unit(1, "lines"),
        legend.position="top",
        legend.justification="right",
        legend.margin=margin(-10,0,-5,-10),
        legend.box.margin=margin(-10,0,-5,-10))

toptheme <- theme_light() + 
  theme(axis.text.y = element_markdown(),
        axis.title = element_blank(),
        axis.ticks = element_blank(),
        strip.background =element_rect(fill="white"),
        strip.text = element_text(colour = "black"),
        panel.spacing = unit(1, "lines"))
middletheme <- theme_light() + 
  theme(axis.text.y = element_markdown(),
        axis.title = element_blank(),
        axis.ticks = element_blank(),
        strip.background =element_blank(),
        strip.text = element_blank(),
        panel.spacing = unit(1, "lines"))
bottomtheme <- theme_light() + 
  theme(axis.title.y=element_blank(),
        axis.text.y=element_markdown(),
        axis.ticks = element_blank(),
        strip.background =element_blank(),
        strip.text = element_blank(),
        panel.spacing = unit(1, "lines"))
csize <- 3
CBdark2.3class <- c('#d95f02', '#7570b3','#1b9e77')
```

## absolute abundance across species

Only randomseed=2 is plotted

```{r}
plotting <- results.densities %>% 
  filter(is.na(hgtclass) | (hgtclass == 2 & randomseed == 2)) %>% 
  unnest(density) %>% 
  mutate(density.per.ha = density*10) %>% 
  left_join(types, by = 'nin') %>% 
  mutate(hgtclassname = if_else(is.na(hgtclass), 
                                'height \u2265 30 cm',
                                'height \u2265 100 cm'))
perspecies <- plotting %>% 
  group_by(hgtclass, species) %>% 
  summarise(daa = sum(daa), n = sum(n)) %>% 
  mutate(species.per.ha = n*10/daa, .keep = "unused")
surveyedarea <- plotting %>% 
  filter(is.na(hgtclass)) %>% 
  group_by(nin) %>% 
  summarise(total.ha = sum(daa)/10) %>% 
  left_join(types, by = join_by(nin)) %>%
  mutate(name = paste0(name, ', ', round(total.ha), ' ha'))
plotting <- plotting %>% 
  left_join(perspecies, by = join_by(species, hgtclass)) %>% 
  select(-name) %>% 
  left_join(surveyedarea, by = join_by(nin))
```

```{r}
d <- plotting %>% 
  arrange(desc(total.ha)) %>% 
  mutate(name = fct_reorder(name, total.ha, first, .na_rm = TRUE))

g <- ggplot(d, aes(x = density.per.ha, y = name)) +
  geom_blank() +
  geom_vline(aes(xintercept = species.per.ha, color = species), linewidth = 1,
             alpha = 0.5, linetype = 1) +
  geom_linerange(aes(xmin=species.per.ha, xmax=density.per.ha, color = species),
                 position = position_dodge(width = 0.5),
                 linetype = '11', alpha = 0.5) +
  geom_point(aes(color = species), size = csize, shape = 19,
             position = position_dodge(width = 0.5)) +
  scale_color_manual(values = c("P.sitchensis-lutzii" = CBdark2.3class[1],
                                "P.abies" = CBdark2.3class[2],
                                "Larix" = CBdark2.3class[3])) +
  coord_cartesian(xlim = c(1, 100)) +
  scale_x_log10(breaks = c(1, 10, 100),
                labels = c('1', '10', '100')) +
  facet_grid(cols = vars(hgtclassname)) +
  labs(title = "<span style = 'color:#d95f02;'>*Picea sitchensis* / 
       *Picea* \u00D7 *lutzii*</span>, 
       <span style = 'color:#7570b3;'>*Picea abies*</span>, 
       <span style = 'color:#1b9e77;'>*Larix* spp.</span>",
       x = "surveyed wildling abundance (ha<sup>-1</sup>)") +
  topbottomtheme +
  theme(legend.position="none")
```

```{r}
ggsave(g, filename= 'abundance.svg', 
       path = here('ms', 'figures'),
       scale = 1.4, width = 210, height = 297/2, units = 'mm', dpi = 300)
```

## relative abundances by species

Only randomseed=2 is plotted

```{r}
plotting.mod <- results.models %>% 
  filter(is.na(hgtclass) | (hgtclass == 2 & randomseed == 2)) %>% 
  select(-data, -model) %>% 
  unnest(preds) %>% 
  filter(is.na(locality)) %>% 
  mutate(modeled = TRUE)
plotting.dens <- results.densities %>% 
  filter(is.na(hgtclass) | (hgtclass == 2 & randomseed == 2)) %>% 
  unnest(density) %>% 
  filter(!(n == 0)) %>% 
  mutate(modeled = FALSE) %>% 
  select(-daa, -n, -density)
plotting <- bind_rows(plotting.dens, plotting.mod) %>% 
  left_join(types, by = 'nin') %>% 
  mutate(hgtclassname = if_else(is.na(hgtclass), 'height \u2265 30 cm',
                                'height \u2265 100 cm'),
         pointshape = if_else(modeled, 'susceptibility (modelled)', 
                              'vulnerability (surveyed)'))

layoutheights <- distinct(plotting, species, name) %>% 
  group_by(species) %>% 
  summarize(n = n()) %>% 
  arrange(desc(species)) %>% 
  pull(n)
```

### P.sitchensis-lutzii

```{r}
d <- filter(plotting, species == "P.sitchensis-lutzii", randomseed == 2 | is.na(randomseed)) %>% 
  arrange(desc(hgtclass), desc(modeled)) %>% 
  mutate(name.br = fct_reorder(name.br, fit, first))
g <- ggplot(d, aes(x = fit, y = name.br, shape = pointshape)) + 
  geom_blank() +
  geom_hline(yintercept = which(levels(d$name.br) == 'forest'), linetype = "dotted") +
  geom_linerange(mapping = aes(xmin = ci.l, xmax = ci.u), color = CBdark2.3class[1], linewidth = 1) +
  geom_point(color = CBdark2.3class[1], size = csize) +
  scale_x_log10(labels = scales::trans_format("log10", scales::math_format(10^.x))) +
  scale_shape_manual(values = c(20, 1)) +
  coord_cartesian(xlim = c(1e-4, 1e2)) +
  facet_grid(cols = vars(hgtclassname)) +
  labs(title = "<span style = 'color:#d95f02;'>*Picea sitchensis* / *Picea* \u00D7 *lutzii*</span>",
       x = 'relative wildling abundance') +
  guides(shape = guide_legend(ncol = 2, title = element_blank(), reverse = TRUE)) +
  topbottomtheme
```

```{r}
ggsave(g, filename= 'Ps.svg', 
       path = here('ms', 'figures', 'susceptibility'),
       scale = 1.4, width = 210, height = layoutheights[1]*4+20, units = 'mm', dpi = 300)
```

### P.abies

```{r}
d <- filter(plotting, species == "P.abies", randomseed == 2 | is.na(randomseed)) %>% 
  arrange(desc(hgtclass), desc(modeled)) %>% 
  mutate(name.br = fct_reorder(name.br, fit, first))
g <- ggplot(d, aes(x = fit, y = name.br, shape = pointshape)) + 
  geom_blank() +
  geom_hline(yintercept = which(levels(d$name.br) == 'forest'), linetype = "dotted") +
  geom_linerange(mapping = aes(xmin = ci.l, xmax = ci.u), color = CBdark2.3class[2], linewidth = 1) +
  geom_point(color = CBdark2.3class[2], size = csize) +
  scale_x_log10(labels = scales::trans_format("log10", scales::math_format(10^.x))) +
  scale_shape_manual(values = c(20, 1)) +
  coord_cartesian(xlim = c(1e-5, 1e1)) +
  facet_grid(cols = vars(hgtclassname)) +
  labs(title = "<span style = 'color:#7570b3;'>*Picea abies*</span>",
       x = 'relative wildling abundance') +
  guides(shape = guide_legend(ncol = 2, title = element_blank(), reverse = TRUE)) +
  topbottomtheme
```

```{r}
ggsave(g, filename= 'Pa.svg', 
       path = here('ms', 'figures', 'susceptibility'),
       scale = 1.4, width = 210, height = layoutheights[2]*4+20, units = 'mm', dpi = 300)
```

### Larix

```{r}
d <- filter(plotting, species == "Larix", randomseed == 2 | is.na(randomseed)) %>% 
  arrange(desc(hgtclass), desc(modeled)) %>% 
  mutate(name.br = fct_reorder(name.br, fit, first))
g <- ggplot(d, aes(x = fit, y = name.br, shape = pointshape)) + 
  geom_blank() +
  geom_hline(yintercept = which(levels(d$name.br) == 'forest'), linetype = "dotted") +
  geom_linerange(mapping = aes(xmin = ci.l, xmax = ci.u), color = CBdark2.3class[3], linewidth = 1) +
  geom_point(color = CBdark2.3class[3], size = csize) +
  scale_x_log10(labels = scales::trans_format("log10", scales::math_format(10^.x))) +
  scale_shape_manual(values = c(20, 1)) +
  coord_cartesian(xlim = c(1e-3, 1e3)) +
  facet_grid(cols = vars(hgtclassname)) +
  labs(title = "<span style = 'color:#1b9e77;'>*Larix* spp.</span>",
       x = 'relative wildling abundance') +
  guides(shape = guide_legend(ncol = 2, title = element_blank(), reverse = TRUE)) +
  topbottomtheme
```

```{r}
ggsave(g, filename= 'L.svg', 
       path = here('ms', 'figures', 'susceptibility'),
       scale = 1.4, width = 210, height = layoutheights[3]*4+20, units = 'mm', dpi = 300)
```

## sensitivity to random height assignment seed

```{r}
plotting <- results.models %>% 
  select(-data, -model) %>% 
  unnest(preds) %>% 
  filter(is.na(locality)) %>% 
  left_join(types, by = "nin") %>% 
  mutate(hgtclassname = if_else(is.na(hgtclass), 'height \u2265 30 cm',
                                'height \u2265 100 cm'))

d <- filter(plotting, species == "P.sitchensis-lutzii", !is.na(hgtclass)) %>% 
  arrange(match(randomseed, c(2, 1, 3)), desc(hgtclass)) %>% 
  mutate(name = fct_reorder(name, fit, first))
g1 <- ggplot(d, aes(x = fit, y = name, group = randomseed)) + 
  geom_blank() +
  geom_hline(yintercept = which(levels(d$name) == 'forest'), linetype = "dotted") +
  geom_linerange(mapping = aes(xmin = ci.l, xmax = ci.u), color = CBdark2.3class[1], size = 0.5,
                 position = position_dodge(width = 0.5)) +
  geom_point(color = CBdark2.3class[1], size = 1, position = position_dodge(width = 0.5)) +
  scale_x_log10(labels = scales::trans_format("log10", scales::math_format(10^.x))) +
  coord_cartesian(xlim = c(1e-4, 1e2)) +
  facet_grid(cols = vars(hgtclassname)) +
  toptheme

d <- filter(plotting, species == "P.abies", !is.na(hgtclass)) %>% 
  arrange(match(randomseed, c(2, 1, 3)), desc(hgtclass)) %>%
  mutate(name = fct_reorder(name, fit, first))
g2 <- ggplot(d, aes(x = fit, y = name, group = randomseed)) + 
  geom_blank() +
  geom_hline(yintercept = which(levels(d$name) == 'forest'), linetype = "dotted") +
  geom_linerange(mapping = aes(xmin = ci.l, xmax = ci.u), color = CBdark2.3class[2], size = 0.5,
                 position = position_dodge(width = 0.5)) +
  geom_point(color = CBdark2.3class[2], size = 1, position = position_dodge(width = 0.5)) +
  scale_x_log10(labels = scales::trans_format("log10", scales::math_format(10^.x))) +
  coord_cartesian(xlim = c(1e-5, 1e1)) +
  facet_grid(cols = vars(hgtclassname)) +
  middletheme

d <- filter(plotting, species == "Larix", !is.na(hgtclass)) %>% 
  arrange(match(randomseed, c(2, 1, 3)), desc(hgtclass)) %>% 
  mutate(name = fct_reorder(name, fit, first))
g3 <- ggplot(d, aes(x = fit, y = name, group = randomseed)) + 
  geom_blank() +
  geom_hline(yintercept = which(levels(d$name) == 'forest'), linetype = "dotted") +
  geom_linerange(mapping = aes(xmin = ci.l, xmax = ci.u), color = CBdark2.3class[3], size = 0.5, 
                 position = position_dodge(width = 0.5)) +
  geom_point(color = CBdark2.3class[3], size = 1, position = position_dodge(width = 0.5)) +
  scale_x_log10(labels = scales::trans_format("log10", scales::math_format(10^.x))) +
  coord_cartesian(xlim = c(1e-3, 1e3)) +
  facet_grid(cols = vars(hgtclassname)) +
  labs(x = 'relative wildling abundance') +
  bottomtheme

layoutheights <- distinct(plotting, species, name) %>% 
  group_by(species) %>% 
  summarize(n = n()) %>% 
  arrange(desc(species)) %>% 
  pull(n)
g123 <- g1 + g2 + g3 + 
  plot_layout(nrow = 3, heights = layoutheights)
```

```{r}
ggsave(g123, filename= 'susceptibility-sensitivity-randomseed.svg', 
       path = here('ms', 'figures'),
       scale = 1, width = 210, height = 297, units = 'mm', dpi = 300)
```

## sensitivity to offsetting seed rain

```{r}
results.sensitivity.offsetting <- results.sensitivity.offsetting %>% 
  rename(fit.end = fit)
plotting <- results.models %>% 
  filter(is.na(randomseed)) %>% 
  select(species, preds) %>% 
  unnest(preds) %>%
  filter(is.na(locality)) %>%
  select(species, nin, fit) %>% 
  left_join(results.sensitivity.offsetting, by = c('species', 'nin')) %>% 
  mutate(diff = fit.end - fit) %>% 
  left_join(types, by = "nin") %>% 
  mutate(hgtclassname = 'height \u2265 30 cm')

 
d <- filter(plotting, species == "P.sitchensis-lutzii") %>% 
  mutate(name = fct_reorder(name, fit, last))
g1 <- ggplot(d, aes(x = fit, y = name)) + 
  geom_blank() +
  geom_hline(yintercept = which(levels(d$name) == 'forest'), linetype = "dotted") +
  geom_segment(data = filter(d, diff != 0), 
               mapping = aes(yend = name, x = fit, xend = fit.end), 
               color = CBdark2.3class[1], linewidth = 2, 
               lineend = 'round', linejoin = 'mitre',
               arrow = arrow(length = unit(0.001, "npc"), angle = 45, type = "closed")) +
  geom_point(data = filter(d, diff == 0),
             mapping = aes(x = fit), color = CBdark2.3class[1], size = 2) +  
  scale_x_log10(labels = scales::trans_format("log10", scales::math_format(10^.x))) +
  coord_cartesian(xlim = c(1e-4, 1e2)) +
  facet_grid(cols = vars(hgtclassname)) +
  toptheme

d <- filter(plotting, species == "P.abies") %>% 
  mutate(name = fct_reorder(name, fit, last))
g2 <- ggplot(d, aes(x = fit, y = name)) + 
  geom_blank() +
  geom_hline(yintercept = which(levels(d$name) == 'forest'), linetype = "dotted") +
  geom_segment(data = filter(d, diff != 0), 
               mapping = aes(yend = name, x = fit, xend = fit.end), 
               color = CBdark2.3class[2], linewidth = 2, 
               lineend = 'round', linejoin = 'mitre',
               arrow = arrow(length = unit(0.001, "npc"), angle = 45, type = "closed")) +
  geom_point(data = filter(d, diff == 0),
             mapping = aes(x = fit), color = CBdark2.3class[2], size = 2) +  
  scale_x_log10(labels = scales::trans_format("log10", scales::math_format(10^.x))) +
  coord_cartesian(xlim = c(1e-5, 1e1)) +
  middletheme

d <- filter(plotting, species == "Larix") %>% 
  mutate(name = fct_reorder(name, fit, last))
g3 <- ggplot(d, aes(x = fit, y = name)) + 
  geom_blank() +
  geom_hline(yintercept = which(levels(d$name) == 'forest'), linetype = "dotted") +
  geom_segment(data = filter(d, diff != 0), 
               mapping = aes(yend = name, x = fit, xend = fit.end), 
               color = CBdark2.3class[3], linewidth = 2, 
               lineend = 'round', linejoin = 'mitre',
               arrow = arrow(length = unit(0.001, "npc"), angle = 45, type = "closed")) +
  geom_point(data = filter(d, diff == 0),
             mapping = aes(x = fit), color = CBdark2.3class[3], size = 2) +  
  scale_x_log10(labels = scales::trans_format("log10", scales::math_format(10^.x))) +
  coord_cartesian(xlim = c(1e-3, 1e3)) +
  labs(x = 'relative wildling abundance') +
  bottomtheme

layoutheights <- distinct(plotting, species, name) %>% 
  group_by(species) %>% 
  summarize(n = n()) %>% 
  arrange(desc(species)) %>% 
  pull(n)
g123 <- g1 + g2 + g3 + 
  plot_layout(nrow = 3, heights = layoutheights)
```

```{r}
ggsave(g123, filename= 'susceptibility-sensitivity-offset.svg', 
       path = here('ms', 'figures'),
       scale = 1, width = 210, height = 297, units = 'mm', dpi = 300)
```

# sessionInfo 

```{r sessionInfo}
sessionInfo()
```