---
title: 'Accounting for confounders reveals that ecosystems vary considerably in susceptibility to alien conifer establishment'
author:
- Julien Vollering^[julienvollering@gmail.com]
- Siri Lie Olsen
- Olav Skarpaas
- Leif Appelgren
- Magni Olsen Kyrkjeeide
- Anders Often
- Jakob Sandven
- Odd Stabbetorp
- Øyvind Sørhuus
bibliography: invasibility-ms.bib
link-citations: true
output:
  bookdown::pdf_document2
documentclass: article
csl: ecology.csl
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, tidy=TRUE, tidy.opts = list(width.cutoff=60))
options(knitr.graphics.auto_pdf = TRUE)
library(tidyverse)
library(here)
library(knitr)
library(kableExtra)
```

# ORCIDs

Julien Vollering: <https://orcid.org/0000-0002-7409-2898>

Siri Lie Olsen: <https://orcid.org/0000-0002-4443-8261>

Olav Skarpaas: <https://orcid.org/0000-0001-9727-1672>

Magni Olsen Kyrkjeeide: <https://orcid.org/0000-0002-7454-3652> 

# Abstract

1. Plantations of alien conifer species are common worldwide, and set to become even more prevalent in coming decades. The rate at which their offspring colonize surroundings varies among plantations, and the reasons for this variation are often unclear. To minimize the spread of so-called "wildlings" and the burden of conifer plantations on native ecosystems, managers need to know which ecosystems are most and least susceptible.
2. We compared how likely wildlings are to establish across a wide range of ecosystems, focusing on four groups of alien conifer species planted in Norway. We used data from detailed surveys around 82 plantations to model the relationship between ecosystem type and wildling abundance --- accounting for seed rain, climate, and other sources of variation between sites. Furthermore, we tested whether differences in susceptibility between individual ecosystem types could be generalized based on broad, shared characteristics.
3. We found that ecosystem susceptibility, modeled as relative establishment likelihood, varied more than the raw densities of wildlings in these ecosystems. Moreover, the rank order of ecosystems by susceptibility differed from their rank order by wildling density. For all species groups, relative establishment likelihoods spanned several orders of magnitude between the most and least susceptible ecosystems.
4. The four groups of conifer species showed somewhat similar patterns of establishment likelihood across ecosystem types, with intensively farmed ecosystems repeatedly among the least susceptible. We found that ecosystems characterized by destabilizing disturbance tended to be most susceptible, but broad ecosystem characteristics did not clarify patterns of susceptibility much, neither within nor across species groups. 
5. *Synthesis and applications* Differences in wildling establishment between ecosystems can be exploited to keep alien conifers within plantation boundaries. Plantations hemmed in by agriculture or other unsusceptible ecosystems will result in relatively few wildlings, while plantations near landslides and other susceptible ecosystems will need monitoring and control. Managers should be aware that the observed density of wildlings in a given ecosystem may not reflect its relative susceptibility, because variation in seed rain, climate, and site characteristics obscures the relationship between ecosystem type and wildling establishment.

# Keywords

alien species, conifer, establishment likelihood, disturbance, invasibility, plantation, recruitment, seed dispersal, Wald Analytical Long-distance Dispersal (WALD)

# Introduction

Plantations of alien conifers are widespread globally, and offspring from these plantations frequently establish in surrounding areas [@richardsonConifersInvasiveAliens2004].
In many instances these naturalized offspring, or wildlings, harm biodiversity and other values, so controlling their spread is prudent.
In particular, plantations contribute to the presence of alien conifers in protected areas, and generate substantial costs as a result [@mcconnachieEstimatingEffectPlantations2015].
Controlling wildlings protects plantation surroundings and prevents secondary, potentially invasive spread.
Accordingly, guidelines for sustainable use of alien trees stress that containing them to the areas set aside for their cultivation is fundamental to good forestry practice [@brunduGlobalGuidelinesSustainable2020].

How many wildlings are present and how far they are from the plantation varies a lot from site to site, even among conspecific plantations of similar age [@fernandesWhatDrivesEucalyptus2018; @nygaardSpreadIntroducedSitka2017], which makes it harder to predict and manage their spread.
To begin to understand this variation, we need to consider both dispersal and establishment (where establishment comprises germination and survival).
These jointly generate patterns of wildling abundance and spatial distribution.
Dispersal at a given site is affected by conditions governing a species' dispersal syndrome --- for instance wind exposure and site topography, in the case of a wind-dispersed species.
Likewise, establishment is affected by biotic and abiotic conditions where seeds arrive.
In theory, conditions inhibiting either process may be exploited to suppress wildling spread.

Reducing establishment is of particular interest because it directly regulates wildling abundance. 
The conditions affecting establishment are also generally easier to manipulate than those affecting dispersal, either directly through intervention or indirectly through site selection.
The question, then, becomes: how can we identify establishment-inhibiting conditions in a manner applicable to plantation management?

In Norway --- and similarly in other countries --- there exists a national land mapping classification system that sorts variation in local ecological conditions [@halvorsenSystematicsEcodiversityEcoSyst2020].
It is a working tool for land management in Norway, with government backing.
The system builds on the continuum concept [@austinContinuumConceptOrdination1985], aiming for reproducible and value-neutral classification of ecosystems by rule-based discretization of species turnover along important environmental gradients [@halvorsenSystematicsEcodiversityEcoSyst2020].
As a result, it encapsulates in its ecosystem types (hereafter: "ecosystems") much of the variation that is most likely to regulate wildling germination and survival --- in competition, nutrient availability, disturbance, and the like [@richardsonNaturalizationIntroducedPlants2012].
It also identifies broad similarities between ecosystems, which might be used to tease out generic trends in establishment likelihood.

To estimate how likely wildlings are to establish in particular ecosystems based on observations of their abundance around a sample of plantations, we must account for (1) seed dispersal, and (2) sources of variation in establishment beside ecosystem.
For example, low establishment likelihood in an ecosystem frequently located close to plantations can be masked by copious seed rain [@rougetInferringProcessPattern2003].
Likewise, an ecosystem may appear to promote establishment if it tends to co-occur with climatic conditions that support germination.
Unconfounded establishment likelihoods can then be used to predict how vulnerable the surroundings of an unobserved plantation are, based on its composition of ecosystems.
  
Determining which ecosystems are most vulnerable so that interventions can be prioritized objectively is among the most urgent objectives for invasion science [@pysekScientistsWarningInvasive2020].
Plantations of wind-dispersed, alien conifers present an opportunity to assess ecosystem invasibility (to these species), because we can estimate ecosystem exposure (seed rain) directly, rather than by proxy [@catfordQuantifyingLevelsBiological2012]. 
We examine plantations in Norway to investigate the following questions:  

1. In which ecosystems are wildlings of alien conifers most and least likely to establish?
2. How do estimates of establishment likelihood differ from observed wildling densities?
3. Can overarching characteristics of ecosystems be used to generalize patterns of wildling establishment?

# Methods

## Field data

We registered wildlings and ecosystems around 82 plantations across Norway, comprising four groups of alien conifers (hereafter: "species"; fig. \@ref(fig:sites-map)).
The sample contained (1) forty-two sites with Sitka spruce (*Picea sitchensis*) or its fertile hybrid, Lutz spruce (*Picea* x *lutzii*), (2) nineteen with Norway spruce (*Picea abies*), (3) fifteen with undetermined larches (*Larix* spp.), and (4) six with lodgepole pine (*Pinus contorta*).
Note that the plantations of Norway spruce were located in parts of Norway where it is not considered native.
We selected and delineated plantations manually --- using aerial imagery --- aiming for plantations that were isolated from other plantations of the same species.
Only plantations with cone-bearing individuals were included in the sample.
We collected field data from each plantation in one of six field campaigns during the period 2016-2019 [reported in @olsenKartleggingAvKortdistansespredning2016; @appelgrenKartleggingAvKortdistansespredning2017; @kyrkjeeideKartleggingAvKortdistansespredning2017; @appelgrenKartleggingAvKortdistansespredning2018; @olsenKartleggingAvKortdistansespredning2019; @sandvenKartleggingAvKortdistansespredning2019].

```{r sites-map, fig.cap='Locations of the 82 plantations in the data set.'}
knitr::include_graphics('figures/sites-map.pdf')
```

In the field, wildlings and ecosystems were mapped comprehensively within a 500x500 m plot centered around the plantation of interest.
Except in the 2016 field campaign, we also mapped as polygons any additional plantations of the same species within a larger 2x2 km plot.
We used GPS to register the point-positions of all wildlings over 30 cm in height, recording a single position for groups of wildlings occurring with less than 5 m between them.
A few exceptionally dense groups of wildlings were mapped by registering polygons instead of points and estimating the number of individuals by transect counts.
Concurrently, we registered polygons for all terrestrial ecosystems, following the Nature in Norway classification system [version 2.0 or 2.1, @halvorsenNaturNorgeNiN2015].
The Nature in Norway system is the national standard for land cover mapping and provides full spatial coverage (i.e. any location and any kind of land cover is assignable to an ecosystem).
We mapped ecosystems at a scale of 1:5000, which implies that all polygons with a size over 250 m^2^ were registered [@brynVeilederKartleggingAv2015].
Regularly patterned occurrence of more than one ecosystem in polygons smaller than the minimum size were registered as so-called mosaic polygons.

Some non-spatial data were also collected for each of the central plantations. 
We measured the height of a representative individual in the plantation by clinometer.
We also estimated the age of the plantation at the time of the field campaign, either by contacting land owners or municipal officials, or by counting growth rings.
Details for all 82 plantations are provided in the Appendix (table \@ref(tab:sites-table)).

## Seed dispersal

To account for the influence of seed dispersal on wildling abundance, we needed estimates of the spatial distribution of seed rain within the 500x500 m field plots.
Because conspecific plantations were sometimes located nearby the central plantation, we considered all plantations within a 1 km radius to be potential seed sources.
We used two models of seed dispersal to derive two different estimates of relative seed rain in space.

The first model was an empirical seed dispersal kernel, with parameter estimates generalized from multiple data sets.
Specifically, we selected from Bullock et al. [-@bullockSynthesisEmpiricalPlant2017] the kernel that performed best for wind-adapted seeds from 5-15 m tall trees (an Exponential Power function).
Seventy-two of the 82 plantations in our data set matched this height range better than a taller range with a different empirical kernel.

The second model was an anisotropic (directional) implementation of the Wald Analytical Long-distance Dispersal [WALD, @katulMechanisticAnalyticalModels2005] model, following Skarpaas et al. [-@skarpaasDispersalPatternsDispersal2007].
We parameterized the model with: site- and season-specific wind vectors retrieved from meteorological data sets, wind turbulence estimated from local ecosystem composition, seed release height based on plantation height, and species-specific seed terminal velocities from literature.

We transformed field-mapped polygons of seed sources into hexagonally gridded point sources, with a density of 0.1 m^-2^ for the first model and 0.01 m^-2^ for the second model (to reduce computation time). 
Then we applied our two dispersal models to estimate the distribution of seed rain from all point sources in a grid of 10 m cells.
We chose this cell size to be similar to the smallest allowed ecosystem polygon.

A full description of our implementation of the WALD model and additional details about seed source polygons are given in the [Appendix].

## Establishment likelihood

For our analysis of establishment likelihood, we rasterized wildling occurrences and ecosystems to the same 10 m grid as the seed dispersal models (fig. \@ref(fig:site-example)). 
Rather than assigning a single ecosystem to each grid cell, we applied fuzzy logic and assigned ecosystems in proportion to their areal coverage of the cell.
In other words, each ecosystem was rendered as a separate raster variable with values in the range [0,1].
By allowing cell composition to be mixed, we tried to capture ecotones in the model and avoid using mapped polygon boundaries overconfidently.
Area covered by mosaic polygons was divided evenly among the constituent types.
We excluded the "tree plantation" ecosystem because some of the field campaigns did not register wildlings when they occurred in this ecosystem.

```{r site-example, out.width='90%', fig.cap='An illustration of one of the 82 plantation sites: Meaasen. The background map shows the surroundings of the plantation, and the 500x500 m plot is overlaid with an aerial photograph. The middle row of panels represents the data as registered in the field (nature type polygons not shown). The top and bottom rows of panels represent selected variables for the 500x500 m plot, as used in the regression model (with a spatial grain of 10 m). Grid cells without data are either seed sources (corresponding to the polygons shown) or tree plantations of other species (such as one along the road).'}
knitr::include_graphics('figures/site-example/site-example.pdf')
```

We used a directed acyclic graph (DAG) to diagram causal relationships among the factors we expected to influence the number of wildlings per cell (fig. \@ref(fig:DAG)).
In the DAG, the unmeasured, proximate causes of wildling abundance --- establishment likelihood and seed rain --- are descendants of variables that we could observe or model.
We included an effect of elevation from plantation on seed rain because neither of our models of seed dispersal account for uneven terrain.
To estimate the causal influence of ecosystem on wildlings, the DAG dictated that we adjust for site, climate, elevation from plantation, and relative seed rain in our regression model [@textorRobustCausalInference2016; @mcelreathHauntedDAGCausal2020].
We did not need to adjust for plantation age (with site in the model), but we included it as a covariate anyway because it could reduce the unexplained variance associated with the random effect of site, and because our assumed causal structure would allow us to interpret its coefficient as an (unconfounded) total effect on wildling abundance [@westreichTableFallacyPresenting2013].
Climate entered the regression model as mean annual temperature (Bio1) and precipitation of the coldest quarter (Bio19), at 30-arcsecond resolution, from CHELSA data [@kargerClimatologiesHighResolution2017].
We chose these variables because they showed the strongest correlations with Norwegian vegetation zones and sections, respectively [@bakkestuenSteplessModelsRegional2008].
In the lodgepole pine model we used only Bio19 because the two variables were highly correlated ($\rho$ = 0.98).
Elevation from plantation was taken with respect to the highest point of the central plantation, from digital elevation models at 1 or 10 m resolution (Norwegian Mapping Authority).
Lastly, plantation surroundings are likely to experience large stochastic variation in annual seed rain, because seed release will coincide with different wind conditions from year to year.
Therefore we also included in our model the first order interaction between relative seed rain and plantation age.

```{r DAG, fig.cap="A directed acyclical graph showing the causal relationships motivating our statistical model of ecosystems' effects on wildling establishment. Red variables causally affect both the ecosystem type and wildling establishment, blue variables causally affect only wildling establishment, and grey variables are unobserved. Green arrows show the causal pathway of interest."}
knitr::include_graphics(here('ms', 'figures', 'dagitty-model.pdf'))
```

For all species, wildling abundance showed a high frequency of zeros that was underestimated by the best fitting negative binomial distribution.
Accordingly, we applied zero-inflated, negative binomial (ZINB) generalized linear models.
These regard zeros as the mixed product of a binomial process as well as a (conditional) count process.
We modeled the binomial process as dependent on plantation age and site. 
We expected that younger plantations would exhibit more cells without wildlings than predicted under a constant rate of establishment, because of their infertile juvenile period.
We also expected the frequency of zeros to vary with site, because our field work documented that some land owners had occasionally made efforts to remove wildlings.
Excess zeros that arose in these ways would therefore not bias our estimates of establishment likelihood [@blasco-morenoWhatDoesZero2019].
To summarize, we modeled:

\begin{equation}
\begin{aligned}
wildlings_{ij} &\sim ZINB(\pi_{ij}, \mu_{ij}, \phi) \\
logit(\pi_{ij}) &= PlantationAge_{ij} + Site_{ij} \\
log(\mu_{ij}) &= \sum\limits_{k=1}^K EcosystemType_{ijk} + Site_{i} + Bio1_{ij} + Bio19_{ij} + \\
&RelativeElevation_{ij} + RelativeSeedRain_{ij} + PlantationAge_{ij} + \\
&RelativeSeedRain_{ij} \cdot PlantationAge_{ij}  \\
Site_{i} &\sim Normal(0, \sigma^2) \\
\end{aligned}
\end{equation}

where $\pi$ is the probability of a zero from the binomial process, while $\mu$ and $\phi$ are the mean and dispersion of the negative binomial distribution, respectively [eq. 11.21 in @zuurZeroTruncatedZeroInflatedModels2009]. 
Subscripts $i$, $j$, and $k$ index sites, cells, and ecosystems.

We standardized variables (except ecosystems, which were [0,1]) and fitted models with the glmmTMB package [version 1.0, @brooksGlmmTMBBalancesSpeed2017] in R [version 3.6, @rcoreteamLanguageEnvironmentStatistical2020].
For each species we dropped ecosystems that were completely free of wildlings (to avoid model convergence issues stemming from complete separation).
We fitted parallel models for our two seed rain variables (derived from empirical or WALD dispersal kernels) and selected the model with the better AIC.
To catch problems with our model specification, we looked for deviation from uniformity in quantile-scaled, simulated residuals, using the DHARMa package [version 0.2.7, @hartigDHARMaResidualDiagnostics2020].
We also ran DHARMa's tests for residual over/underdispersion and zero-inflation.
The Norway spruce model showed residual underdispersion, so we replaced the negative binomial distribution with the generalized Poisson distribution, which solved the issue [@brooksStatisticalModelingPatterns2019].
Furthermore, the lodgepole pine model did not converge as specified.
To achieve convergence we dropped site from the binomial submodel, reasoning that its variance was difficult to estimate from only six sites.
Finally, establishment likelihoods for ecosystems were calculated as predictions from the conditional part of the zero-inflation model (holding covariates at their mean values), scaled by the value predicted for "forests".

To check how the covariates in our models swayed estimates of establishment likelihood, we calculated the observed density of wildlings in each ecosystem (abundance/area).
To facilitate comparison, we derived these densities from the same gridded data set used to train the models, tallying wildlings in proportion to the ecosystem composition of the grid cells they occupied. 
For example, a cell occupied by one wildling and half its area covered by a given ecosystem would tally 0.5 wildling for that ecosystem.

To test whether higher-level characteristics of ecosystems can be used to generalize patterns of vulnerability, we aggregated ecosystems by their category (terrestrial or wetland) or structuring process (anthropogenic, natural or no disturbance), as defined in the Nature in Norway system (Appendix, table \@ref(tab:types-table)).
We then refitted our (otherwise unchanged) models twice --- once with each of these stratifications replacing ecosystems --- and obtained estimates of relative establishment likelihood for each category and structuring process.

# Results

Estimated establishment likelihoods varied by up to five orders of magnitude (fig. \@ref(fig:establishment)).
The pattern of establishment likelihood among ecosystems was modestly similar between species, with positive rank correlations in four of six species pairs (tab. \@ref(tab:correlation-table)).

```{r establishment, fig.cap="Relative densities (unfilled points) and relative establishment likelihoods (filled points) of four alien conifer species groups in various ecosystem types, using 'forest' as the reference level. Zero density is plotted at the lower limit of the x-axis. Estimates of relative establishment likelihood are shown with 95 % confidence intervals. Grey point clouds depict relative establishment likelihoods for individual sites in our data set. The order of ecosystem types along the y-axis is determined by a confidence-weighted mean of their percentile-ranked establishment likelihoods within species, such that types with consistently high establishment likelihood across species are at the top. Wetland ecosystem types are shown in italic font. Ecosystem types without wildlings and covering less than one hectare across all sites for that species are not displayed. Photos licensed CC BY 4.0 Rune Halvorsen."}
knitr::include_graphics(here('ms', 'figures', 'establishment', 'establishment.pdf'))
```

```{r correlation-table}
cortab <- read.csv(here('ms', 'tables', 'spearman-corr.csv'), 
                   row.names = 1, check.names = FALSE)
cortab[lower.tri(cortab, diag = TRUE)] <- ""
cortab <- cortab[-4,-1]
kbl(cortab, booktabs = TRUE,
    caption = "Rank correlations of relative establishment likelihoods in ecosystem types, between pairs of species.") %>%
  kable_styling(full_width = F)
```

Relative establishment likelihoods varied more than the corresponding set of observed densities, in all species (fig. \@ref(fig:establishment)).
Observed densities across all ecosystems ranged 0-211/ha for Sitka spruce, 0-49/ha for Norway spruce, 0-1045/ha for larch, and 0-219/ha for lodgepole pine.
The ranking of ecosystems by establishment likelihood also differed from  their ranking by observed density.
For instance, Sitka/Lutz spruce was observed at higher densities in "artificial substrate" than "boreal heath", but was more likely to establish in "boreal heath".

For all species, the WALD seed dispersal model produced seed rain values that predicted wildling establishment better than the values derived from the empirical seed dispersal kernel (Appendix, table \@ref(tab:model-comparison-table)).
The direct effect of relative elevation on establishment --- not including its effect mediated by ecosystems --- was modest and acted in different directions for different species (Appendix, tables \@ref(tab:Ps)-\@ref(tab:Pc)).
The direct effect of climate on establishment --- not including its effect mediated by ecosystems --- was negligible for Sitka/Lutz spruce and Norway spruce. 
For larch there was a negative direct effect of precipitation (Bio19), with a 50 mm increase in coldest-quarter precipitation estimated to decrease establishment six-fold.
Likewise for lodgepole pine, the model estimated a 175 mm increase in coldest-quarter precipitation to decrease establishment by a factor of 50.
For Sitka/Lutz spruce and lodgepole pine, establishment likelihood varied strongly between sites, such that it swamped much of the variation between ecosystems.

Variation in establishment likelihoods shrank when ecosystems were aggregated (by category or structuring process; fig. \@ref(fig:establishment-hypotheses)).
No more than two orders of magnitude separated the different groups.
Sitka/Lutz spruce and Norway spruce established at higher rates in wetland ecosystems than terrestrial ecosystems, while the opposite was true for larch and lodgepole pine.
Sitka/Lutz spruce and Norway spruce did not generally colonize disturbance-structured ecosystems at higher rates than ecosystems without disturbance structuring.
For larch and lodgepole pine, structuring by natural disturbance and anthropogenic disturbance, respectively, were associated with the highest establishment likelihoods. 
Yet despite larger sample sizes, the mean establishment likelihood in any particular group was rarely clearly distinguishable from that in other groups.

```{r establishment-hypotheses, fig.cap="Relative densities (unfilled points) and relative establishment likelihoods (filled points) of four alien conifer species groups among categories of ecosystem types (top) or structuring processes in ecosystem types (bottom). Estimates of relative establishment likelihood are shown with 95 % confidence intervals. The horizontal dotted lines mark the reference level for the scaling. Strata without wildlings and covering less than one hectare across all sites for that species are not displayed."}
knitr::include_graphics(here('ms', 'figures', 'establishment-hypotheses.pdf'))
```

# Discussion

## Ecosystem vulnerability

In a large database of vegetation plots across Europe, Chytrý et al.[-@chytryHabitatInvasionsAlien2008] found that alien plants as a group are consistently found at low rates in mires and heaths, and high rates in arable, man-made, and coastal ecosystems.
The conifer species we examined --- perhaps with the exception of lodgepole pine --- do not conform to these broader trends in ecosystem invasibility.
Lodgepole pine establishment is relatively well studied, and our results are consistent with this literature.
"Bare rock" harbors very few competitors and showed highest lodgepole pine establishment of all ecosystems [@despainDispersalEcologyLodgepole2001], "coastal heath" (a short shrubland) facilitated more establishment than grassland-like ecosystems [@taylorDriversPlantInvasion2016], all ecosystems with canopy cover were among those with lowest establishment [@langdonPinusContortaInvasion2010; @taylorDriversPlantInvasion2016], and ecosystems with high establishment were frequently structured by anthropogenic disturbance [@richardsonPineInvasionsSouthern1994].

It is difficult to evaluate our ecosystem vulnerability results against the recruitment patterns that have been described for the three other species.
For instance, Sitka spruce grows poorly under moisture stress and tolerates flooding well [@petersonEcologyManagementSitka1997], which might account for why it was twice as likely to establish in wetland ecosystems as in terrestrial ecosystems.
Yet it also established in "open shallow-soil ground" with very high likelihood, despite this ecosystem's characteristically dry soil. 
This illustrates the trouble with deriving predictions for management units such as ecosystems from generalized statements about species autecology; should we expect few wildlings in "open shallow-soil ground" because it is dry, or many because it provides ample light and reduced competition [@petersonEcologyManagementSitka1997]?
Furthermore, ecosystems that would seem inhospitable based on their overall characteristics may actually contain many localized opportunities for establishment, because seedling mortality is strongly regulated by microsites [@macekLifeDeathPicea2017].
From this perspective, our estimates of establishment likelihood measure the density of suitable microsites in a given ecosystem.

The breadth in establishment likelihood suggests that differences between ecosystems deserve careful consideration when managing wildling spread.
This knowledge can be applied in at least two ways.
First, as a preventative measure, we recommend siting new plantations where surrounded by high proportions of ecosystems with low establishment likelihood.
In particular, "arable fields" harbor very few wildlings of any species and are common near existing plantations, so picking sites hemmed in by this kind of agricultural land should be both effective and feasible.
This would probably reduce the rate of wildling establishment by orders of magnitude, even if long distance dispersal might preclude complete containment [@albertLanduseChangeSubalpine2008]. 
In some cases it may also be desirable to alter ecosystems adjacent to existing plantations to prevent (further) spread, for example by intensifying mowing regimes to promote the appearance of "agriculturally improved grassland".
Second, as a reactive measure, we recommend allocating resources for monitoring and control in proportion to relative ecosystem vulnerability.
Prioritizing ecosystems that are highly vulnerable and also rare (e.g. "open active landslide") is especially likely to be cost-efficient. 

The establishment patterns we quantify probably hold, more or less, beyond Norway [@chytryHabitatInvasionsAlien2008].
From a manager's perspective, we expect that the ecosystems we report may translate well to equivalent types in similar classification systems, because the Nature in Norway classification is rule-based and aims for observer neutrality. 
At the same time, we urge caution in extending our establishment estimates to ecosystems that are only broadly similar, because we found that similar types frequently showed markedly different vulnerability (e.g. Norway spruce in "agriculturally improved grassland with semi-natural character" vs. "agriculturally improved grassland").

An observational study like ours informs management of long-lived, naturalized species more directly than experimental studies, because longer time frames are examined.
It measures long-term survival --- often the quantity of interest --- under a wide range of natural conditions experienced by the wildlings.
In contrast, seeding experiments generally observe only the youngest life stages, and the factors controlling individual success differ at later life stages [@dovciakSeedRainEnvironmental2008].
For example, Sitka spruce appears more likely to germinate in disturbed soil [@vikaneInvasionCallunaHeath2013], but less likely to survive there [@petersonEcologyManagementSitka1997].
On the other hand, experiments might be more useful when observed wildling spread is not representative of patterns in the wider landscape (e.g. for species expanding from a single point of introduction).

## Estimating vulnerability

Confounders of the relationship between ecosystem and wildling abundance caused observed wildling density to underestimate and sometimes completely mischaracterize differences in establishment likelihood between ecosystems.
The underestimation implies that confounding variables counteracted differences in ecosystem establishment in our sample --- for instance that vulnerable ecosystems were concentrated at sites predicted to show limited establishment.
We caution, therefore, that direct inference from observations of wildling abundance misguides intuition about relative ecosystem vulnerability.
For example, the observed density of Sitka spruce wildlings was about equal in "boreal heath" and "sand dunes", but we estimate that establishment likelihood is actually about 100 times larger in "boreal heath".

The effect size of WALD-modeled relative seed rain was estimated close to one for all species except Norway spruce, which indicates that spatial patterns of wildling abundance around plantations were captured satisfactorily by this variable.
We take this as evidence that the WALD model describes seed dispersal well in our system, but recognize that survivorship bias prevents rigorous assessment of this relationship with our data.
For example, the empirical seed dispersal model we tested generally showed a steeper decline in seed rain away from seed sources than the WALD model, so negative density dependence in seedling survival could skew the spatial distribution of wildling abundance towards the WALD model.
That this important covariate was modeled and not measured is a limitation of our method, and it makes the establishment likelihoods we estimate less certain.
Nevertheless, the mechanistic nature of the WALD model makes us more confident in its estimates across species and sites than we would be in a purely phenomenological model [@bullockAllDispersalFunctions2018]. 
Relative elevation's inconsistent effects on wildling abundance suggests that either terrain does not strongly affect seed dispersal or --- more likely --- vertical distance alone does a poor job of representing its effect.
In any case, there is no rule of thumb for management that wildlings tend to move up or down slopes.

We reiterate that our models do not estimate the total causal effect of climate on wildling abundance, because they set aside climate's influence on ecosystems [@westreichTableFallacyPresenting2013].
Therefore, we interpret the estimated climate effects with respect to physiological constraints within a given ecosystem. 
The lack of a direct climate effect on Sitka/Lutz spruce wildling abundance is consistent with Sitka spruce's wide climatic tolerance relative to climatic variation in Norway [@petersonEcologyManagementSitka1997; @volleringBunchingBackgroundBetters2019, Appendix F].
Norway spruce seedling recruitment has previously been found to increase towards the wetter end of Norwegian climate [@tingstadTemperaturePrecipitationBiotic2015], but most of our Norway spruce sites circumscribed a narrow part of that range.
We are not confident that lodgepole pine responds strongly to precipitation, as estimated, because the sample contained only six sites from two climates.

A curious feature of our results that needs more research is the large amount of unexplained variation in Sitka/Lutz spruce and lodgepole pine establishment likelihood between sites.
Our ability to predict these species' magnitude of spread at a specific site, relative to other sites, is still limited. 
Ecosystem comparisons can nevertheless guide management.
Bianchi et al. [-@bianchiMethodsPredictingSitka2019] struggled to predict regeneration density within Sitka spruce plantations from bare ground cover, moss cover, plantation age, and plantation density.
The inadequacy of plantation density as a predictor in this context suggests that the unexplained site-level variation we observed was not caused by differences in the densities of seed sources (which we assumed to be constant).
Alternative explanations could include: (1) property owners removing wildlings at some sites, or (2) differing demographic characteristics among plantations [@taylorDriversPlantInvasion2016], potentially as a result of provenance. 

## Generalizing vulnerability

The overarching characteristics that we used to aggregate ecosystems did a poor job of generalizing differences in vulnerability.
That is, groups of ecosystems belonging to the same hydrological category, or structured by the same form of disturbance, showed heterogeneous establishment likelihoods.
We note that slightly different sets of ecosystems comprised the groups for each species, depending on their presence in the data.
These differences in ecosystem composition help explain why the patterns of aggregated establishment likelihood varied between species. 
This constraint hinders species comparisons but underlines our main takeaway from these results --- that the vulnerability of an individual ecosystem frequently diverges from those it is classified with.
Thus, we did not find any broad commonalities between vulnerable ecosystems that could help guide wildling management where data are scarce. 

Within species, we urge careful interpretation of the comparisons among ecosystem categories and structuring processes.
Many areas where conifer establishment is nearly impossible, like paved surfaces and annually plowed fields, count as terrestrial and anthropogenically disturbed, which lowers the relative establishment likelihood of these two groups.
Our results do not imply, for example, that any particular anthropogenic disturbance event will decrease establishment likelihood of Sitka/Lutz spruce relative to an ecosystem's prior state (indeed, Vikane et al. [-@vikaneInvasionCallunaHeath2013] show that burning in coastal heathland increases Sitka spruce establishment).
Rather, we find that ecosystems structured by anthropogenic disturbance, on the whole, are no more vulnerable to Sitka/Lutz spruce wildlings than other ecosystems.

# Conclusions

Wildling spread from plantations is a growing problem [@richardsonConifersInvasiveAliens2004] and will probably worsen with recent pushes to increase tree planting worldwide [@brunduGlobalGuidelinesSustainable2020]. 
Meanwhile, remotely sensed and survey data are making detailed and accurate maps of ecosystems increasingly available over large extents [@horvathDistributionModellingVegetation2019], which presents opportunities to manage wildling spread more efficiently [@buckleySlowingPineInvasion2005].
Specifically, differences in ecosystem vulnerability can be leveraged to 
reduce the rate of wildling establishment (potentially by orders of magnitude) through deliberate site selection for new plantations or targeted interventions around existing plantations.
However, managers should be cautious judging ecosystem vulnerability based on descriptions of the species' autecology, observed wildling densities, or generalizations about vulnerability across ecosystems.

One of main novelties of this study is that we inferred vulnerability/invasibility using mechanistically reconstructed, spatial estimates of seed rain.
Scientists studying invasibility at national and continental scales have already recognized the importance of normalizing observed levels of invasion by a spatially explicit estimate of exposure [i.e. propagule pressure; @chytrySeparatingHabitatInvasibility2008]. 
However, many studies quantifying ecosystem invasibility have not been able adjust for propagule pressure, typically because it is impossible to reconstruct the underlying dispersal history [@catfordQuantifyingLevelsBiological2012].
We found that accounting for seed rain and other confounders of the relationship between ecosystems and wildling abundance underlined variation in ecosystem invasibility.

# Authors' contributions

JV, SLO, and OSk conceived the ideas and designed methodology; SLO, LA, MOK, AO, JS, OSt, and ØS collected the data; JV analysed the data and led the writing of the manuscript. All authors contributed critically to the drafts and gave final approval for publication.

# Acknowledgements

We thank Honorata Gajda, Knut Børge Strøm, Heidi Elin Myklebost, Jon Hagelin, Vigdis Frivoll, Sina Thu Randulff, Roy Mangersnes, Bjarne Homnes Oddane, Rune Søyland, Solbjørg Engen Torvik, Toralf Tysse, Craig Jackson, Miene-Marie Gastinger, Ola Westby Aamodt, and Anders Ringstad for their help in collecting and/or collating the field data.

# Appendix

```{r table-setup, include = FALSE}
options(knitr.table.format = "latex")
```

The WALD model takes the form of an inverse Gaussian distribution whose mean ($\mu$) and shape ($\lambda$) parameters are calculated from physical characteristics of the dispersal system:

\begin{equation}
\mu  = \frac{HU}{F}
\end{equation}
\begin{equation}
\lambda = \left(\frac{H}{\sigma}\right)^2
\end{equation}

where $H$ is the seed release height, $U$ is the mean horizontal wind velocity between $H$ and the ground, $F$ is the terminal velocity of the seed, and $\sigma$ is a wind turbulence parameter.
We set $H$ to the height of the central plantation, estimated $U$ from a computed vertical wind profile, obtained $F$ from literature, and calculated $\sigma$ from an equation for turbulent flow as a function of vegetation height [eq. A4 in @skarpaasDispersalPatternsDispersal2007].
We parameterized separate WALD models for 20º sectors around each seed source, to make seed dispersal anisotropic (directional).
In each sector we estimated mean vegetation height based on the composition of mapped ecosystem types (Appendix, table \@ref(tab:types-table)).
Simultaneously, we randomly sampled 100 wind velocities in the direction of the sector during the species' dispersal season.
The 100 resulting WALD kernels produced the seed probability density in the sector, and individual sectors were weighed by the frequency of corresponding wind directions (again, during the species' dispersal season).
The wind data were obtained either from the nearest weather station (MET Norway), a 2.5 km resolution interpolated hindcast covering southern Norway [@haakenstad15yearHighResolution2017], or a 10 km resolution hindcast covering all of Norway [@reistadHighresolutionHindcastWind2011; @haakenstadNORA10EIRevisedRegional2020]. We used weather station data if the station was less than 2.5 or 10 km away (depending on hindcast coverage), or else the highest resolution hindcast.

```{r recreate traits-table, include=FALSE}
traits <- tribble(
  ~`species group`, ~`seed terminal velocity`, ~`dispersal season`, ~references,
  "Larix spp.", "1.0 m/s", "Dec - May", "2, 3",
  "Picea abies", "0.58 m/s", "Nov - May", "2, 4",
  "Pinus contorta", "0.82 m/s", "Sep - Dec", "2",
  "Picea sitchensis / lutzii", "0.94 m/s", "Oct - Feb", "1, 2")
```

```{r traits-table}
kable(traits, longtable = TRUE, booktabs = TRUE, escape = FALSE, 
      caption = "Dispersal traits") %>% 
  column_spec(1, italic = TRUE) %>%
  footnote(general = c("1. Harris, A. S. Sitka spruce. in Silvics of North America: 1. Conifers (eds. Burns, R. M. & Honkala, B. H.) vol. 2 513–529 (U.S. Department of Agriculture, Forest Service, 1990).",
                       "2. Sandvik, H. Kunnskapsstatus for spredning og effekter av fremmede bartrær på biologisk mangfold. (2012).",
                       "3. Sullivan, J. Larix decidua. Fire Effects Information System, U.S. Department of Agriculture, Forest Service, Rocky Mountain Research Station,  Fire Sciences Laboratory https://www.fs.fed.us/database/feis/plants/tree/lardec/all.html (1994).",
                       "4. Sullivan, J. Picea abies. Fire Effects Information System, U.S. Department of Agriculture, Forest Service, Rocky Mountain Research Station,  Fire Sciences Laboratory https://www.fs.fed.us/database/feis/plants/tree/picabi/all.html (1994)."), 
            general_title = "References: ",
           threeparttable = TRUE)
```

Some of the seed source polygons we registered in the field had distinctive features that we accounted for as follows.
Seed source polygons for which the species of interest only made up a fraction of the plantation composition [e.g. in @olsenKartleggingAvKortdistansespredning2019] were used with their point source density adjusted accordingly.
For example, a plantation identified as composed of Sitka spruce and Norway spruce was assigned a seed source point density half that of a pure Sitka spruce plantation.
Likewise, 'mixed forest' plantations [e.g. in @appelgrenKartleggingAvKortdistansespredning2018] were assigned 0.1 times the seed source point density of a pure plantation.  
Seed source polygons identified as logged [e.g. in @appelgrenKartleggingAvKortdistansespredning2018] were included as seed sources only if we could confirm that they were logged no earlier the decade prior to mapping, using time series of aerial photos.

```{r load sites-table, include=FALSE}
locs <- read_csv(here('output', 'localities-modified.csv')) # see 'wildlings-NiN.Rmd'
locs <- locs %>% 
  select(year.registered, reference, species, 
         locality, utm33.easting, utm33.northing,
         height.source, age, bio01, bio19,
         height.interpolated, age.interpolated) %>% 
  arrange(year.registered, reference, species) %>% 
  select(-year.registered)
# Formatting
locs <- locs %>% 
  mutate(species = case_when(
    species == "P.contorta" ~ "Pinus contorta",
    species == "P.sitchensis-lutzii" ~ "Picea sitchensis / lutzii",
    species == "P.abies" ~ "Picea abies",
    species == "Larix" ~ "Larix spp.")) %>% 
  mutate(locality = paste(toupper(substring(locality, 1, 1)), 
                          substring(locality, 2), sep = ""),
             .before = utm33.easting) %>%
  mutate_at(c("utm33.easting","utm33.northing"), round) %>% 
  mutate_at(c("bio01","bio19"), ~ round(./10, digits = 2)) %>% 
  rename(`species group` = species, site = locality, easting = utm33.easting, 
         northing = utm33.northing, height = height.source)
# Footnotes
names(locs)[8] <- paste0(names(locs)[8], footnote_marker_alphabet(1))
names(locs)[9] <- paste0(names(locs)[9], footnote_marker_alphabet(2))
locs <- locs %>% 
  mutate_at(c("height","age"), as.character) %>% 
  mutate(height = case_when(
    height.interpolated ~ paste0(height, footnote_marker_symbol(1)),
    TRUE ~ height)) %>% 
  mutate(age = case_when(
    age.interpolated ~ paste0(age, footnote_marker_symbol(2)),
    TRUE ~ age)) %>% 
  select(-height.interpolated, -age.interpolated)
```

```{r sites-table}
kbl(locs, longtable = TRUE, align = "l", booktabs = TRUE, escape = FALSE, 
    caption = "\\label{tab:sites-table}Plantations") %>% 
  kable_styling(latex_options =c("repeat_header","striped")) %>% 
  column_spec(2, italic = TRUE) %>%
  footnote(general = "Easting and Northing are given for UTM zone 33N. Height is given in meters and age in years.",
           alphabet = c("mean annual temperature (°C)",
                        "precipitation in coldest quarter (cm)"),
           symbol = c("interpolated as the mean height of other plantations of the same species, inversely weighted by difference in age",
                      "interpolated as the mean age of other plantations of the same species in the same region")) %>% 
  landscape()
```

```{r make types table, include=FALSE}
pairing <- read_csv(here('data','codetype-pairing.csv'))
types <- read_csv(here('data', 'types.csv'))
typestab <- inner_join(types, pairing) %>%
  filter(!grepl(pattern = '.', x = .data$code, fixed = TRUE)) %>% 
  select(type = name, code, category, structuring, `vegetation height` = vegheight)
# Footnotes
names(typestab)[5] <- paste0(names(typestab)[5], footnote_marker_alphabet(1))
```

```{r types-table}
options(knitr.kable.NA = '')
kbl(typestab, longtable = TRUE, booktabs = TRUE, escape = FALSE, 
    caption = "\\label{tab:types-table}Ecosystem types") %>% 
  kable_styling(latex_options =c("repeat_header", "striped")) %>% 
  footnote(alphabet = "approximate vegetation heights (meters) are used only to estimate wind turbulence") %>% 
  landscape()
```

```{r model-comparison-table}
options(knitr.kable.NA = 'NA')
mcomptab <- read.csv(here('ms', 'tables', 'modelcomp.csv'), check.names = FALSE)
kbl(mcomptab, booktabs = TRUE,
    caption = "Comparison of models with different seed dispersal estimates") %>%
  kable_styling(full_width = F) %>% 
  collapse_rows(columns = 1, latex_hline = "major", valign = "middle")
```

```{r load model summaries, message=FALSE}
msumms <- list.files(here('ms', 'tables'), pattern = "summary", full.names = TRUE) %>% 
  set_names(~ tools::file_path_sans_ext(basename(.))) %>% 
  map(read_csv) %>% 
  map(~ mutate(., Term = if_else(Term == "locality", "site", Term)))
```

```{r summaries-tables}
kbl(msumms$`Ps-modelsummary`, longtable = TRUE, booktabs = TRUE, align = 'clrrrr', 
    caption = "\\label{tab:Ps}Model summary for Picea sitchensis / lutzii. The conditional submodel is glmmTMB's nbinom2 family with dispersion parameter phi = 0.253.") %>%
  kable_styling(font_size = 6) %>% 
  add_header_above(c(" " = 2, "Fixed effects" = 2, "Random effects" = 2)) %>% 
  column_spec(1, bold = T) %>% 
  collapse_rows(columns = 1, valign = "top")

kable(msumms$`Pa-modelsummary`, longtable = TRUE, booktabs = TRUE, align = 'clrrrr', caption = "\\label{tab:Pa}Model summary for Picea abies. The conditional submodel is glmmTMB's genpois family with dispersion parameter phi = 2.59.") %>%
  kable_styling(font_size = 6) %>% 
  add_header_above(c(" " = 2, "Fixed effects" = 2, "Random effects" = 2)) %>% 
  column_spec(1, bold = T) %>% 
  collapse_rows(columns = 1, valign = "top")

kable(msumms$`L-modelsummary`, longtable = TRUE, booktabs = TRUE, align = 'clrrrr', caption = "\\label{tab:L}Model summary for Larix. The conditional submodel is glmmTMB's nbinom2 family with dispersion parameter phi = 0.261.") %>%
  kable_styling(font_size = 6) %>% 
  add_header_above(c(" " = 2, "Fixed effects" = 2, "Random effects" = 2)) %>% 
  column_spec(1, bold = T) %>% 
  collapse_rows(columns = 1, valign = "top")

kable(msumms$`Pc-modelsummary`, longtable = TRUE, booktabs = TRUE, align = 'clrrrr', caption = "\\label{tab:Pc}Model summary for Pinus contorta. The conditional submodel is glmmTMB's nbinom2 family with dispersion parameter phi = 0.0269.") %>%
  kable_styling(font_size = 6) %>% 
  add_header_above(c(" " = 2, "Fixed effects" = 2, "Random effects" = 2)) %>% 
  column_spec(1, bold = T) %>% 
  collapse_rows(columns = 1, valign = "top")
```

# References

```{r sessionInfo}
capture.output(sessionInfo(), file="sessionInfo.txt")
```
