---
title: 'Alien conifer spread in Norway'
author:
  - Julien Vollering:
      email: julienvollering@gmail.com
      institute: [IMN, NHM]
      correspondence: true
  - name: Et Al
    institute: NHM
institute:
  - IMN: Department of Environmental Sciences, Western Norway University of Applied Sciences, Sogndal, Norway
  - NHM: Department of Research and Collections, Natural History Museum, University of Oslo, Oslo, Norway
bibliography: invasibility-ms.bib
output: 
  bookdown::html_document2:
    pandoc_args:
      - '--lua-filter=lua-filters/scholarly-metadata.lua'
      - '--lua-filter=lua-filters/author-info-blocks.lua'
always_allow_html: true
editor_options:
  chunk_output_type: console
link-citations: yes
csl: ecology.csl
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, tidy=TRUE, tidy.opts = list(width.cutoff=60))
options(knitr.graphics.auto_pdf = TRUE)
library(tidyverse)
library(here)
library(knitr)
library(kableExtra)
```

# ORCIDs

Julien Vollering: <https://orcid.org/0000-0002-7409-2898>

# Running head

Alien conifers in Norway

# Abstract


# Keywords


# Introduction

Plantations of alien conifers are widespread in Norway, as they are in many countries, and offspring from these plantations frequently establish in surrounding areas [@richardsonConifersInvasiveAliens2004].
Around some plantations these naturalized offspring --- or wildlings --- are abundant and distant from the plantation boundary.
Around others there are few or no wildlings [@nygaardNaturligSpredningAv1999].

Depending on how rapidly wildlings spread, their expansion beyond the plantation may be considered invasive. 
Implications for integrity of the ecosystem, especially important for areas of high value, and potentially affecting ecosystem services.
For example, biodiversity, coastal heathland, landscape changes.
As a result, managers are interested in limiting spread from plantations of alien species (compared to natural regeneration of mixed stand forestry)

Norway has only two native conifer species that cover large areas: Scots pine, *Pinus sylvestris* and Norway spruce, *Picea abies*).
As a result, 20th century efforts to increase forest cover relied heavily on alien species [@oyenBiomassPotentialSelected2017]. 

'Invasive' normally applied to species, but differences among plantations of the same species are also observed.

The degree of naturalized spread observed surrounding plantations is highly variable.
Specifically, the number of naturalized individuals, their location with respect to the plantation, and their rate of growth varies. 
Many factors may be hypothesized to affect the degree of naturalized spread: the species in the plantation, the seed production of the plantation, the dispersal capacity of the the seeds, and the establishment rate of the seeds in the surroundings.  

It is important to consider at what spatial scale variability in natural recruitment occurs. This may differ between species. If most variation occurs at the coarsest spatial scale, drivers that also vary at course scales, such as climate, are likely to be important. If variation between plantations is greater than variation between regions, climate and other coarse-scale factors are probably less important, and drivers that vary at the plantation scale are likely important. If more variation occurs around single plantations than between plantations, site characteristics (e.g. surrounding ecosystem types) are likely the most important governing factors. 

Fertile plantations produce highest propagule pressure on their immediate surroundings, because their seeds are (primarily?) wind-dispersed.
Extensive
LDD less important

Managers are interested in limiting or eliminating spread from plantations, especially into areas with high ecological value (protected areas?).
Therefore, the key process that we examine is naturalized spread, which can be broken down into the processes of seed arrival and then germination, growth, and survival (assuming that the seed is viable when it arrives).
Seed arrival depends on dispersal from source populations, and is a fundamentally spatial process. 
Meanwhile, germination, growth, and survival (hereafter: "establishment") depend on the conditions at the arrival location.
To better understand which factors govern naturalized spread, we need to know how seeds are dispersed from plantations, and which landing site conditions promote or inhibit establishment. 

Therefore, the number of wildlings in the surroundings of a plantation is a function of the seed shadow from the plantation and the environmental characteristics of the proximal area.
Note that the seed shadow is not purely a function of plantation characteristics; plantation surroundings also affect the seed shadow.  
For example, lower elevations may trap more seeds.

# Methods

## Approach

We compared the frequency with which four groups of alien conifer species establish in various ecosystem types, using observational data.
Specifically, we used maps of wildlings and ecosystem types around the plantations from which the wildlings likely originated.
The groups of species (hereafter: "species") we examined were: (1) Sitka spruce (*Picea sitchensis*) together with its fertile hybrid, Lutz spruce (*Picea* x *lutzii*), (2) undetermined larches (*Larix* spp.), (3) Norway spruce (*Picea abies*), and (4) lodgepole pine (*Pinus contorta*).
We analysed the data in two ways, to answer distinct questions about patterns of wildling establishment.

Our first analysis tallied of the number of wildlings of each species per unit area of each ecosystem type.
In this way, we summarized the state of wildling establishment from alien conifer plantations under prevailing conditions.
In the tallying analysis, variation in the density of wildlings is partly attributable to differences between species and between ecosystem types, and partly to contingent properties of our sample, such as the spatial distribution of ecosystem types around plantations.
For example, if a particular ecosystem type tended to occur close to plantations, it would likely show higher wildling density than other ecosystem types even if the species were equally likely to establish there.
The motivation for this approach was descriptive: to break down the realized spread of wildlings from the specific plantations in the data set.

Our second analysis was designed to isolate the influence of ecosystem types on the establishment probability of each species.
To do so, we defined establishment probability (E) as the ratio of wildlings (W) to deposited seeds (S) in a given area, or:

$$W_{xy} 	= S_{xy}\; * E_{xy}$$

where xy denotes a unit area.
Note that this simple conceptual model mirrors our basic mechanistic understanding of the spread process: a seed (S) disperses from its source to a new location, and then the conditions at that location (E) determine the likelihood that the seed establishes as a wildling (W).
Guided by this equality, we specified a statistical model for the relationship between wildlings and ecosystem types:
$S_{xy}$ was assumed to be a function of the plantation's age and its probability of dispersing seeds to that area, while $E_{xy}$ was assumed to be a function of climate and ecosystem type.
Plantation age was included in the model because the observed wildlings were not an annual cohort but rather accrued over the lifetime of the plantation.

## Field data

We observed wildlings and ecosystem types around 63 plantations across Norway (fig. \@ref(fig:localities-map)).
The sample comprised thirty-six localities with Sitka/Lutz spruce, twelve with larch, nine with Norway spruce, and six with lodgepole pine.
Note that the plantations of Norway spruce were located in parts of Norway where it is not considered native [@aarrestadForeignNorwaySpruce2014].
We selected and delineated plantations manually --- using aerial imagery --- aiming for plantations that were isolated from other plantations of the same species.
Furthermore, only plantations with cone-bearing individuals were included in the sample.
We collected the field data from each plantation in one of five field campaigns during the period 2016-2019 [reported in @olsenKartleggingAvKortdistansespredning2016; @appelgrenKartleggingAvKortdistansespredning2017; @kyrkjeeideKartleggingAvKortdistansespredning2017; @appelgrenKartleggingAvKortdistansespredning2018; @olsenKartleggingAvKortdistansespredning2019].

```{r localities-map, fig.cap='Locations across Norway of the 63 plantations in the data set.'}
knitr::include_graphics('figures/localities-map.pdf')
```

In the field, wildlings and ecosystem types were mapped comprehensively within a 500x500 m plot centered around the plantation of interest.
Except in the 2016 field campaign, we also mapped as polygons any additional plantations of the same species within a larger 2x2 km plot.
We used GPS to register the point-positions of all wildlings over 30 cm in height, recording a single position for groups of wildlings occurring with less than 5 m between them.
A few exceptionally dense groups of wildlings were mapped by registering polygons instead of points and estimating the number of individuals by transect counts.
Concurrently, we registered polygons for all terrestrial ecosystem types following the Nature in Norway classification system [version 2.0 or 2.1, @halvorsenNaturNorgeNiN2015].
The Nature in Norway system is the national standard for land cover mapping and provides full spatial coverage (i.e. any location and any kind of land cover is assignable to an ecosystem type).
It builds on the continuum concept in ecology, aiming for reproducible and value-neutral classification of ecosystems by rule-based discretization of species turnover along important environmental gradients [@austinContinuumConceptOrdination1985; @NatureTypesNorway2018; @whittakerGradientAnalysisVegetation1967].
We mapped ecosystem types at a scale of 1:5000, which implies that all polygons with a size over 250 m^2^ were registered [@brynVeilederKartleggingAv2015].

Some non-spatial data were also collected for each plantation. 
We measured the height of a representative individual in the plantation by clinometer.
We also estimated the age of the plantation at the time of the field campaign, either by contacting land owners or municipal officials, or by counting growth rings.
A table of the 63 plantations and their accompanying non-spatial data is provided in Appendix [A1].

## Seed dispersal

For our second analysis --- isolating the effect of ecosystem type on establishment probability --- we needed estimates of the spatial distribution of seed deposition within the 500x500 m field plots.
Because conspecific plantations were sometimes located nearby the central plantation, we considered all plantations within a 1 km radius to be potential seed sources.
We used two models of seed dispersal to derive two different estimates of seed probability density.

The first model was an empirical seed dispersal location kernel [@nathanDispersalKernels2012], with parameter estimates generalized from multiple data sets.
Specifically, we selected from Bullock et al. [-@bullockSynthesisEmpiricalPlant2017] the kernel that performed best for wind-adapted seeds from 5-15 m tall trees (an Exponential Power function).
Fifty-seven of the 63 plantations in our data set matched this height range better than a taller height range with a different empirical kernel.

The second model was mechanistic, using local wind velocity and direction to parameterize the Wald Analytical Long-distance Dispersal [WALD, @katulMechanisticAnalyticalModels2005] model in different directions from seed sources [following @skarpaasDispersalPatternsDispersal2007].
The WALD model takes the form of an inverse Gaussian distribution whose mean ($\mu$) and shape ($\lambda$) parameters are calculated from physical characteristics of the dispersal system:
$$\mu  = \frac{HU}{F}$$
$$\lambda = \left(\frac{H}{\sigma}\right)^2$$
where $H$ is the seed release height, $U$ is the mean horizontal wind velocity between $H$ and the ground, $F$ is the terminal velocity of the seed, and $\sigma$ is a wind turbulence parameter.
We set $H$ to the height of the central plantation, estimated $U$ from a computed vertical wind profile, obtained $F$ from literature (Appendix [A2]), and calculated $\sigma$ from an equation for turbulent flow as a function of vegetation height [@skarpaasDispersalPatternsDispersal2007, eq. A4].
We parameterized separate WALD models for 20º sectors around each seed source, to make seed dispersal anisotropic (directional).
In each sector we estimated vegetation height based on the composition of mapped ecosystem types (Appendix [A3]).
Simulataneously, we randomly sampled 100 wind velocities in the direction of the sector during the species' dispersal season (Appendix [A2]), using data from a high resolution atmospheric reanalysis [@haakenstad15yearHighResolution2017] or from the nearest weather station (MET Norway).
The 100 resulting WALD kernels produced the seed probability density in the sector, and individual sectors were weighed by the frequency of corresponding wind directions (again, restricting winds to the species' dispersal season).

Both models of seed probability density were expressed on a grid with a grain of 10 m, so dispersal kernels were applied to the distance from seed source to grid cell center. 
Field-mapped polygons of seed sources were transformed into hexagonally gridded point sources, with a density of 0.1 m^-2^ for the first model and 0.01 m^-2^ for the second model (to reduce computation time). 

## Establishment frequency

(fig. \@ref(fig:locality-example))

```{r locality-example, fig.cap='A illustration of data from one of the 63 plantation localities: Meaasen. The background map shows the surroundings of the plantation, and the 500x500 m plot is overlaid with an aerial photograph. The middle row of panels represents the data as registered in the field (NiN polygons not shown). The top and bottom rows of panels represent selected variables as they enter the regression model (with a spatial grain of 10 m).'}
knitr::include_graphics('figures/locality-example/locality-example.pdf')
```

# Results

- Caution against drawing conclusions about the overall invasiveness of species from the comparison of establishment probabilities. The comparison is done for a particular plantation age, under particular climatic conditions, with a particular amount of seed rain --- all of which may vary systematically between species. For example, most lodgepole pine plantations are planted inland but the comparison is under a more oceanic climate, which increases its predicted rate of establishment.

# Discussion

Advantage to observational study (registering established individuals rather than doing seeding plots) is that seeds experience only natural conditions, and it measures long-term survival --- the quantity of interest for management.

Spatial grain limitations: minimum polygon area NiN, wildling group locations. A good reason to use a relatively course observation units (100 m2) and to perform fuzzy coding of ecosystem types.

# Appendix

## A1

```{r load localities-table, include=FALSE}
locs <- read_csv(here('output', 'localities-modified.csv')) # see 'wildlings-NiN.Rmd'
locs <- locs %>% 
  select(year.registered, reference, species, 
         locality, utm33.easting, utm33.northing,
         height.source, age, bio01, bio19,
         height.interpolated, age.interpolated) %>% 
  arrange(year.registered, reference, species) %>% 
  select(-year.registered)
# Formating
locs <- locs %>% 
  mutate(species = case_when(
    species == "P.contorta" ~ "Pinus contorta",
    species == "P.sitchensis-lutzii" ~ "Picea sitchensis / lutzii",
    species == "P.abies" ~ "Picea abies",
    species == "Larix" ~ "Larix spp.")) %>% 
  mutate(locality = paste(toupper(substring(locality, 1, 1)), 
                          substring(locality, 2), sep = "")) %>%
  mutate_at(c("utm33.easting","utm33.northing"), round) %>% 
  mutate_at(c("bio01","bio19"), ~ round(./10, digits = 2)) %>% 
  rename(`species group` = species, easting = utm33.easting, northing = utm33.northing,
         height = height.source)
# Footnotes
names(locs)[8] <- paste0(names(locs)[8], footnote_marker_alphabet(1))
names(locs)[9] <- paste0(names(locs)[9], footnote_marker_alphabet(2))
locs <- locs %>% 
  mutate_at(c("height","age"), as.character) %>% 
  mutate(height = case_when(
    height.interpolated ~ paste0(height, footnote_marker_symbol(1)),
    TRUE ~ height)) %>% 
  mutate(age = case_when(
    age.interpolated ~ paste0(age, footnote_marker_symbol(2)),
    TRUE ~ age)) %>% 
  select(-height.interpolated, -age.interpolated)
```

```{r localities-table}
kable(locs, escape = FALSE, align = "lllrrrrrr") %>% 
  kable_styling(bootstrap_options = c("striped", "condensed")) %>% 
  column_spec(2, italic = TRUE) %>%
  footnote(general = "Easting and Northing are given for UTM zone 33N. Height is given in meters and age in years.",
           alphabet = c("mean annual temperature (°C)",
                        "precipitation of coldest quarter (cm)"),
           symbol = c("interpolated as the mean height of other plantations of the same species, inversely weighted by difference in age",
                      "interpolated as the mean age of other plantations of the same species in the same region"))
```

## A2

```{r recreate traits-table, include=FALSE}
traits <- tribble(
  ~`species group`, ~`seed terminal velocity`, ~`dispersal season`, ~references,
  "Larix spp.", 1.0, "Dec - May", "2, 3",
  "Picea abies", 0.58, "Nov - May", "2, 4",
  "Pinus contorta", 0.82, "Sep - Dec", "2",
  "Picea sitchensis / lutzii", 0.94, "Oct - Feb", "1, 2")
```

```{r traits-table}
kable(traits, escape = FALSE) %>% 
  kable_styling(bootstrap_options = c("striped", "condensed")) %>% 
  column_spec(1, italic = TRUE) %>%
  footnote(general = c("1. Harris, A. S. Sitka spruce. in Silvics of North America: 1. Conifers (eds. Burns, R. M. & Honkala, B. H.) vol. 2 513–529 (U.S. Department of Agriculture, Forest Service, 1990).",
                       "2. Sandvik, H. Kunnskapsstatus for spredning og effekter av fremmede bartrær på biologisk mangfold. (2012).",
                       "3. Sullivan, J. Larix decidua. Fire Effects Information System, U.S. Department of Agriculture, Forest Service, Rocky Mountain Research Station,  Fire Sciences Laboratory https://www.fs.fed.us/database/feis/plants/tree/lardec/all.html (1994).",
                       "4. Sullivan, J. Picea abies. Fire Effects Information System, U.S. Department of Agriculture, Forest Service, Rocky Mountain Research Station,  Fire Sciences Laboratory https://www.fs.fed.us/database/feis/plants/tree/picabi/all.html (1994)."), 
            general_title = "References: ")
```

## A3

```{r recreate vegheight table, include=FALSE}
types <- read_csv(here('data', 'typenames.csv'))
grouping <- tribble( # from 'locality-WALD.Rmd'
  ~code, ~group, ~vegheight,
  "T1", "T1", 0,
  "T2", "T2", 0.5,
  "T3", "T3", 0.5,
  "T4", "T4", 10,
  "T6", "T6", 0,
  "T12", "T12", 0.5,
  "T13", "T13", 0,
  "T16", "T16", 0.5,
  "T17", "T17", 0,
  "T18", "T18", 0,
  "T21", "T21", 0,
  "T24", "T24", 0.5,
  "T27", "T27", 0,
  "T29", "T29", 0,
  "T30", "T30", 10,
  "T31", "T31", 0.5,
  "T32", "T32", 0.5,
  "T33", "T33", 0.5,
  "T34", "T34", 0.5,
  "T35", "artificial", 0,
  "T36", "T36", 0.5,
  "T37", "artificial", 0,
  "T38", "plantation forest", 10,
  "T39", "artificial", 0,
  "T40", "T40", 0,
  "T41", "T41", 0.5,
  "T42", "T42", 0,
  "T43", "artificial", 0,
  "T44", "T44", 0.5,
  "T45", "T45", 0.5,
  "V1", "V1", 0,
  "V2", "V2", 10,
  "V3", "V3", 0,
  "V4", "V4", 0,
  "V8", "V8", 10,
  "V9", "V9", 0,
  "V10", "V10", 0,
  "V11", "V11", 0,
  "V12", "V12", 0,
  "V13", "V13", 0)  
nin <- left_join(grouping, types, by = c('group' = 'type'))
nin <- select(nin, code, name, vegheight) %>% 
  rename(`vegetation height` = vegheight)
# Footnotes
names(nin)[3] <- paste0(names(nin)[3], footnote_marker_alphabet(1))
```

```{r nin-table}
kable(nin, escape = FALSE) %>% 
  kable_styling(bootstrap_options = c("striped", "condensed")) %>% 
  footnote(alphabet = "approximate vegetation heights (meters) are used only to estimate wind turbulence")
```

# References

```{r sessionInfo}
capture.output(sessionInfo(), file="sessionInfo.txt")
```
