---
title: 'Alien conifer spread in Norway'
author:
  - Julien Vollering:
      email: julienvollering@gmail.com
      institute: [IMN, NHM]
      correspondence: true
  - name: Et Al
    institute: NHM
institute:
  - IMN: Department of Environmental Sciences, Western Norway University of Applied Sciences, Sogndal, Norway
  - NHM: Department of Research and Collections, Natural History Museum, University of Oslo, Oslo, Norway
bibliography: invasibility-ms.bib
output: 
  bookdown::html_document2:
    pandoc_args:
      - '--lua-filter=lua-filters/scholarly-metadata.lua'
      - '--lua-filter=lua-filters/author-info-blocks.lua'
editor_options:
  chunk_output_type: console
link-citations: yes
csl: ecology.csl
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, tidy=TRUE, tidy.opts = list(width.cutoff=60))
options(knitr.graphics.auto_pdf = TRUE)
```

# ORCIDs

Julien Vollering: <https://orcid.org/0000-0002-7409-2898>

# Running head

Alien conifers in Norway

# Abstract


# Keywords


# Introduction

Plantations of alien conifers are widespread in Norway, as they are in many countries, and offspring from these plantations frequently establish in surrounding areas [@richardsonConifersInvasiveAliens2004].
Around some plantations these naturalized offspring --- or wildlings --- are abundant and distant from the plantation boundary.
Around others there are few or no wildlings [@nygaardNaturligSpredningAv1999].

Depending on how rapidly wildlings spread, their expansion beyond the plantation may be considered invasive. 
Implications for integrity of the ecosystem, especially important for areas of high value, and potentially affecting ecosystem services.
For example, biodiversity, coastal heathland, landscape changes.
As a result, managers are interested in limiting spread from plantations of alien species (compared to natural regeneration of mixed stand forestry)

Norway has only two native conifer species that cover large areas: Scots pine, *Pinus sylvestris* and Norway spruce, *Picea abies*).
As a result, 20th century efforts to increase forest cover relied heavily on alien species [@oyenBiomassPotentialSelected2017]. 

'Invasive' normally applied to species, but differences among plantations of the same species are also observed.

The degree of naturalized spread observed surrounding plantations is highly variable.
Specifically, the number of naturalized individuals, their location with respect to the plantation, and their rate of growth varies. 
Many factors may be hypothesized to affect the degree of naturalized spread: the species in the plantation, the seed production of the plantation, the dispersal capacity of the the seeds, and the establishment rate of the seeds in the surroundings.  

It is important to consider at what spatial scale variability in natural recruitment occurs. This may differ between species. If most variation occurs at the coarsest spatial scale, drivers that also vary at course scales, such as climate, are likely to be important. If variation between plantations is greater than variation between regions, climate and other coarse-scale factors are probably less important, and drivers that vary at the plantation scale are likely important. If more variation occurs around single plantations than between plantations, site characteristics (e.g. surrounding nature types) are likely the most important governing factors. 

Fertile plantations produce highest propagule pressure on their immediate surroundings, because their seeds are (primarily?) wind-dispersed.
Extensive
LDD less important

Managers are interested in limiting or eliminating spread from plantations, especially into areas with high ecological value (protected areas?).
Therefore, the key process that we examine is naturalized spread, which can be broken down into the processes of seed arrival and then germination, growth, and survival (assuming that the seed is viable when it arrives).
Seed arrival depends on dispersal from source populations, and is a fundamentally spatial process. 
Meanwhile, germination, growth, and survival (hereafter: "establishment") depend on the conditions at the arrival location.
To better understand which factors govern naturalized spread, we need to know how seeds are dispersed from plantations, and which landing site conditions promote or inhibit establishment. 

Therefore, the number of wildlings in the surroundings of a plantation is a function of the seed shadow from the plantation and the environmental characteristics of the proximal area.
Note that the seed shadow is not purely a function of plantation characteristics; plantation surroundings also affect the seed shadow.  
For example, lower elevations may trap more seeds.

# Methods

## Approach

- Observational
- Mechanisitic motivation for the statistical model
- 4 species/complexes
- complete mapping

We compared the frequency with which four groups of alien conifer species establish in various nature types, using observational data.
Specifically, we used maps of wildlings and nature types around sexually mature plantations from which the wildlings likely originated.
The groups of species (hereafter: "species") we examined were: (1) Sitka spruce (*Picea sitchensis*) together with its hybrid, Lutz spruce (*Picea x lutzii*), (2) undetermined larches (*Larix* spp.), (3) Norway spruce (*Picea abies*), and (4) lodgepole pine (*Pinus contorta*).
We analysed the data in two ways, to answer distinct questions about patterns of wildling establishment.

Our first analysis simply tallied of the number of wildlings of each species per unit area of each nature type.
In this way, we summarized the state of wildling establishment from alien conifer plantations under prevailing conditions.
In the tallying analysis, variation in the density of wildlings is partly attributable to differences between species and between nature types, and partly to contingent factors, such as the spatial distribution of nature types around plantations. 
The motivation for this approach was descriptive: to break down the realized spread of wildlings at the specific localities in the data set.

Our second analysis was designed to isolate the influence of nature types on the establishment probability of each species.
To do so, we defined establishment probability as the ratio of wildlings to deposited seeds in a given area, or:

$$(wildlings)_{xy} 	= (seeds)_{xy}\; * (establishment\; probability)_{xy}$$

where xy denotes a unit area.
Note that this simple conceptual model mirrors our basic mechanistic understanding of the spread process: a seed disperses from its source to a new location, and then the conditions at that location determine the likelihood that the seed establishes as a wildling.
Guided by this equality, we specified a statistical model for the relationship between nature type and wildlings.
The observed number of wildlings in a given area accrued from the 

We estimated the probability density of seed deposition around each plantation by modeling seed dispersal, and assumed that the plantation's total seed production was proportional to .

accounting for the fact that the number of wildlings in a given area could accrue from the time that the source plantation began producing seeds to the time that the observation was made.
In particular, the number of seeds deposited to a given area over the fertile lifetime of the plantation was modeled as proportional to  we made establishment probability a function of nature type and climate, and the number of seeds a function of  
In our observational data, there is also a temporal dimension that is implicit in the model above.
The number of wildlings in a given area could accrue from the time that the source plantation began producing seeds to the time that the observation was made, so the number of deposited seeds must correspond to the same period. 

We also estimated the relative number of seeds arriving at different locations within the mapped areas, by modeling seed dispersal from each plantation. 

Our observational data represent processes that occur over time, so it is worth considering the conceptual model's temporal dimension.
The number of wildlings at a given location accrued from the time that the plantation began producing seeds to the time that the observation was made.
If we knew the absolute number of seeds that arrived during that period, the relationship in the conceptual model would mirror the definition of establishment probability, and could be treated as an equality.
However, we calculated seed arrival as a probability density, which is why the the relationship in the conceptual model is a proportionality.
Similarly, 

$$(wildlings)_{xy,i} 	\propto [(plantation\; age)_{i} * (seed\; probability\; density)_{xy,i}] *  [(climate)_{i} * (nature\; type)_{xy}]$$


Establishment probability was hypothesized to vary primarily by nature type, 

Sustained establishment (includes survival probabilities)

## Field data

We used observations of wildlings and nature types collected around 63 sexually mature conifer plantations, composed of four groups of alien species (hereafter: "species").
The sample comprised: (1) thirty-six plantations of Sitka spruce (*Picea sitchensis*) or its fertile hybrid, Lutz spruce (*Picea* x *lutzii*), (2) twelve plantations of undetermined larches (*Larix* spp.), (3) nine plantations of Norway spruce (*Picea abies*), and (4) six plantations of lodgepole pine (*Pinus contorta*).
Note that the plantations of Norway spruce were located in parts of Norway where it not native [fig. \@ref(fig:localities-map); @aarrestadForeignNorwaySpruce2014].
The plantation localities were selected and visited as a part of one of five field campaigns during the period 2016-2019 [reported in @olsenKartleggingAvKortdistansespredning2016; @appelgrenKartleggingAvKortdistansespredning2017; @kyrkjeeideKartleggingAvKortdistansespredning2017; @appelgrenKartleggingAvKortdistansespredning2018;   @olsenKartleggingAvKortdistansespredning2019]


```{r localities-map, fig.cap='Locations across Norway of the 63 plantations in the data set.'}
knitr::include_graphics('figures/localities-map.pdf')
```

## Modeled seed dispersal

## Statistical modeling of establishment probability

(fig. \@ref(fig:locality-example))

```{r locality-example, fig.cap='A illustration of data from one of the 63 plantation localities: Meaasen. The background map shows the surroundings of the plantation, and the 500x500 m plot is overlaid with an aerial photograph. The middle row of panels represents the data as registered in the field (NiN polygons not shown). The top and bottom rows of panels represent selected variables as they enter the regression model (with a spatial grain of 10 m)'}
knitr::include_graphics('figures/locality-example/locality-example.pdf')
```

# Results

- Caution against drawing conclusions about the overall invasiveness of species from the comparison of establishment probabilities. The comparison is done for a particular plantation age, under particular climatic conditions, with a particular amount of seed rain --- all of which may vary systematically between species. For example, most lodgepole pine plantations are planted inland but the comparison is under a more oceanic climate, which increases its predicted rate of establishment.

# Discussion

Advantage to observational study (registering established individuals rather than doing seeding plots) is that seeds experience only natural conditions, and it measures long-term survival --- the quantity of interest for management.

# References

```{r sessionInfo}
capture.output(sessionInfo(), file="sessionInfo.txt")
```
