---
title: 'Accounting for seed rain reveals large and idiosyncratic differences between ecosystems in vulnerability to alien conifer establishment'
author:
- Julien Vollering^[julienvollering@gmail.com]
- et al.
bibliography: invasibility-ms.bib
output: 
  bookdown::html_document2:
    toc: true
    toc_float: true
always_allow_html: true
editor_options:
  chunk_output_type: console
link-citations: yes
csl: ecology.csl
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, tidy=TRUE, tidy.opts = list(width.cutoff=60))
options(knitr.graphics.auto_pdf = TRUE)
library(tidyverse)
library(here)
library(knitr)
library(kableExtra)
```

# ORCIDs

Julien Vollering: <https://orcid.org/0000-0002-7409-2898>

# Running head

Ecosystem invasibility to alien conifers

# Abstract


# Keywords

conifer, disturbance, invasibility, plantation, seed rain, Wald Analytical Long-distance Dispersal (WALD)

# Introduction

Plantations of alien conifers are widespread globally, and offspring from these plantations frequently establish in surrounding areas [@richardsonConifersInvasiveAliens2004].
In many instances these naturalized offspring, or wildlings, harm biodiversity and other values, so controlling their spread is prudent.
In particular, plantations contribute to the presence of alien conifers in protected areas, and generate substantial costs as a result [@mcconnachieEstimatingEffectPlantations2015].
Controlling wildlings protects plantation surroundings and prevents secondary, potentially invasive spread.
Accordingly, the European "Code of conduct on planted forest and invasive alien trees" stresses that containing alien species to the areas set aside for their cultivation is fundamental to good forestry practice [@brunduPlantedForestsInvasive2016].

How many wildlings are present and how far they are from the plantation varies a lot from site to site, even among conspecific plantations of similar age [@nygaardSpreadIntroducedSitka2017], which makes it harder to predict and manage their spread.
To understand the variation between sites, we need to consider both dispersal and establishment (where establishment comprises germination and survival).
These jointly generate patterns of wildlings in space.
Dispersal at a given site is affected by external conditions governing a species' dispersal syndrome --- for instance wind exposure and site topography, in the case of a wind-dispersed species.
Likewise, establishment is affected by biotic and abiotic conditions where seeds arrive.
In theory, conditions inhibiting either process may be exploited to suppress wildling spread.

Reducing establishment is of particular interest because it regulates wildling abundance directly (by comparison, reducing dispersal can only regulate wildling abundance through negative density dependence, all else being equal). 
The conditions affecting establishment are also generally easier to manipulate than those affecting dispersal, either directly through intervention or indirectly through site selection [@calvino-cancelaInvasivePotentialEucalyptus2013].
The question, then, becomes: how can we identify establishment-inhibiting conditions in a manner applicable to plantation management?

In Norway --- and similarly in other countries --- there exists a national land mapping classification system that sorts variation in local ecological conditions [@NatureTypesNorway2018].
It is a working tool for land management in Norway, with government backing.
The system builds on the continuum concept [@austinContinuumConceptOrdination1985], aiming for reproducible and value-neutral classification of ecosystems by rule-based discretization of species turnover along important environmental gradients [@whittakerGradientAnalysisVegetation1967; @halvorsenNaturNorgeNiN2015].
As a result, it encapsulates in its ecosystem types much of the variation that is most likely to regulate wildling germination and survival --- in competition, nutrient availability, disturbance, and the like [@richardsonNaturalizationIntroducedPlants2012].

We can use ecosystem types as transferable units, to generalize establishment likelihood beyond specific, observed sites.
Importantly, that requires accounting for differences in the amount of seed rain received by the observed ecosystem types [@catfordQuantifyingLevelsBiological2012].
For example, low establishment likelihood in an ecosystem type close to a plantation can be masked by copious seed rain [@rougetInferringProcessPattern2003].
Having quantified establishment likelihood in this way, we can predict how vulnerable the surroundings of an unobserved plantation are, based on its composition of ecosystem types.
  
Determining which ecosystems are most vulnerable so that interventions can be prioritized objectively is among the most urgent objectives for invasion science [@pysekScientistsWarningInvasive2020].
Plantations of wind-dispersed, alien conifers present a golden opportunity to assess ecosystem invasibility (to these species, @hegerPredictingBiologicalInvasions2003), because we can estimate ecosystem exposure (seed rain) directly, rather than by proxy [@catfordQuantifyingLevelsBiological2012]. 
We examine plantations in Norway to investigate the following questions:  

1. In which ecosystem types are wildlings of alien conifers most and least likely to establish?  
2. How do these estimates of establishment likelihood change when variation in seed rain is accounted for?  
3. Can overarching characteristics of ecosystem types be used to generalize patterns of wildling establishment?  

# Methods

## Approach

We compared how frequently four groups of alien conifer species have established in various ecosystem types.
Specifically, we used maps of wildlings and ecosystem types around the plantations from which the wildlings likely originated.
The groups of species (hereafter: "species") we examined were: (1) Sitka spruce (*Picea sitchensis*) together with its fertile hybrid, Lutz spruce (*Picea* x *lutzii*), (2) Norway spruce (*Picea abies*), (3) undetermined larches (*Larix* spp.), and (4) lodgepole pine (*Pinus contorta*). 

For each of the species, we analysed establishment likelihood in two ways.
Our first analysis simply tallied of the number of wildlings per unit area for each ecosystem type.
The motivation for this naive approach was descriptive; we summarized the state of wildling establishment from the plantations in the sample.
Our second analysis was designed to isolate the influence of ecosystem types on establishment likelihood.
To do so, we defined establishment likelihood $E$ as the ratio of wildlings $W$ to seed rain $S$ in a given area, or:

\begin{equation} 
W_{xy} 	= S_{xy} * E_{xy}
(\#eq:estmod)
\end{equation}

where xy denotes a unit area.
This simple conceptual model mirrors our basic mechanistic understanding of the spread process: seeds disperse from their source to a given location ($S_{xy}$), and then the conditions at that location ($E_{xy}$) determine the likelihood that the seed establishes as a wildling ($W_{xy}$).
Guided by this equality, we specified a statistical model for the relationship between wildlings and ecosystem types, conditioning on estimates of $S_{xy}$.


## Field data

We observed wildlings and ecosystem types around 82 plantations across Norway (fig. \@ref(fig:sites-map)).
The sample comprised forty-two sites with Sitka/Lutz spruce, nineteen with Norway spruce, fifteen with larch, and six with lodgepole pine.
Note that the plantations of Norway spruce were located in parts of Norway where it is not considered native [@aarrestadForeignNorwaySpruce2014].
We selected and delineated plantations manually --- using aerial imagery --- aiming for plantations that were isolated from other plantations of the same species.
Furthermore, only plantations with cone-bearing individuals were included in the sample.
We collected the field data from each plantation in one of six field campaigns during the period 2016-2019 [reported in @olsenKartleggingAvKortdistansespredning2016; @appelgrenKartleggingAvKortdistansespredning2017; @kyrkjeeideKartleggingAvKortdistansespredning2017; @appelgrenKartleggingAvKortdistansespredning2018; @olsenKartleggingAvKortdistansespredning2019; @sandvenKartleggingAvKortdistansespredning2019].

```{r sites-map, fig.cap='Locations of the 82 plantations in the data set.'}
knitr::include_graphics('figures/sites-map.svg')
```

In the field, wildlings and ecosystem types were mapped comprehensively within a 500x500 m plot centered around the plantation of interest.
Except in the 2016 field campaign, we also mapped as polygons any additional plantations of the same species within a larger 2x2 km plot.
We used GPS to register the point-positions of all wildlings over 30 cm in height, recording a single position for groups of wildlings occurring with less than 5 m between them.
A few exceptionally dense groups of wildlings were mapped by registering polygons instead of points and estimating the number of individuals by transect counts.
Concurrently, we registered polygons for all terrestrial ecosystem types, following the Nature in Norway classification system [version 2.0 or 2.1, @halvorsenNaturNorgeNiN2015].
The Nature in Norway system is the national standard for land cover mapping and provides full spatial coverage (i.e. any location and any kind of land cover is assignable to an ecosystem type).
We mapped ecosystem types at a scale of 1:5000, which implies that all polygons with a size over 250 m^2^ were registered [@brynVeilederKartleggingAv2015].
Regularly patterned occurrence of more than one ecosystem type in polygons smaller than the minimum size were registered as so-called mosaic polygons.

Some non-spatial data were also collected for each of the central plantations. 
We measured the height of a representative individual in the plantation by clinometer.
We also estimated the age of the plantation at the time of the field campaign, either by contacting land owners or municipal officials, or by counting growth rings.
Details for all 82 plantations are provided in Appendix [A1].

## Seed dispersal

For our second analysis --- isolating the effect of ecosystem type on establishment likelihood --- we needed estimates of the spatial distribution of seed rain within the 500x500 m field plots.
Because conspecific plantations were sometimes located nearby the central plantation, we considered all plantations within a 1 km radius to be potential seed sources.
We used two models of seed dispersal to derive two different estimates of relative seed rain in space.

The first model was an empirical seed dispersal location kernel [@nathanDispersalKernels2012], with parameter estimates generalized from multiple data sets.
Specifically, we selected from Bullock et al. [-@bullockSynthesisEmpiricalPlant2017] the kernel that performed best for wind-adapted seeds from 5-15 m tall trees (an Exponential Power function).
Seventy-two of the 82 plantations in our data set matched this height range better than a taller height range with a different empirical kernel.

The second model was mechanistic, using local wind velocity and direction to parameterize the Wald Analytical Long-distance Dispersal [WALD, @katulMechanisticAnalyticalModels2005] model in different directions from seed sources [following @skarpaasDispersalPatternsDispersal2007].
The WALD model takes the form of an inverse Gaussian distribution whose mean ($\mu$) and shape ($\lambda$) parameters are calculated from physical characteristics of the dispersal system:
$$\mu  = \frac{HU}{F}$$
$$\lambda = \left(\frac{H}{\sigma}\right)^2$$
where $H$ is the seed release height, $U$ is the mean horizontal wind velocity between $H$ and the ground, $F$ is the terminal velocity of the seed, and $\sigma$ is a wind turbulence parameter.
We set $H$ to the height of the central plantation, estimated $U$ from a computed vertical wind profile, obtained $F$ from literature (Appendix [A2]), and calculated $\sigma$ from an equation for turbulent flow as a function of vegetation height [@skarpaasDispersalPatternsDispersal2007, eq. A4].
We parameterized separate WALD models for 20º sectors around each seed source, to make seed dispersal anisotropic (directional).
In each sector we estimated mean vegetation height based on the composition of mapped ecosystem types (Appendix [A3]).
Simulataneously, we randomly sampled 100 wind velocities in the direction of the sector during the species' dispersal season (Appendix [A2]).
The 100 resulting WALD kernels produced the seed probability density in the sector, and individual sectors were weighed by the frequency of corresponding wind directions (again, during the species' dispersal season).
The wind data were obtained either from the nearest weather station (MET Norway), a 2.5 km resolution interpolated hindcast covering southern Norway [@haakenstad15yearHighResolution2017], or a 10 km resolution hindcast covering all of Norway [@reistadHighresolutionHindcastWind2011; @haakenstadNORA10EIRevisedRegional]. We used weather station data if the station was less than 2.5 or 10 km away (depending on hindcast coverage), or else the highest resolution hindcast.

Both estimates of relative seed rain were expressed on a grid with a grain of 10 m, so dispersal kernels were applied to the distance from seed source to grid cell center. 
Field-mapped polygons of seed sources were transformed into hexagonally gridded point sources, with a density of 0.1 m^-2^ for the first model and 0.01 m^-2^ for the second model (to reduce computation time). 
Additional details about how data from different field campaigns were harmonized to model seed dispersal are explained in Appendix [A4].

## Establishment

For our analyses of establishment, we rasterized wildling occurences and ecosystem types to the same 10 m grid as the seed dispersal models (fig. \@ref(fig:site-example)). 
Rather than assigning a single ecosystem type to each grid cell, we applied fuzzy logic and assigned ecosystem types in proportion to their areal coverage of the cell.
In other words, each ecosystem type was rendered as a separate raster variable with values in the range [0,1].
Area covered by mosaic polygons was divided evenly among the constituent types.
We excluded the "tree plantation" ecosystem type from both analyses because some of the field campaigns did not register wildlings occurring within this type.

```{r site-example, fig.cap='An illustration of one of the 82 plantation sites: Meaasen. The background map shows the surroundings of the plantation, and the 500x500 m plot is overlaid with an aerial photograph. The middle row of panels represents the data as registered in the field (NiN polygons not shown). The top and bottom rows of panels represent selected variables as they enter the regression model (with a spatial grain of 10 m). Grid cells without data are either seed sources (corresponding to the polygons shown) or tree plantations of other species (such as one along the road).'}
knitr::include_graphics('figures/site-example/site-example.svg')
```

For our first analysis, we calculated wildling density per ecosystem type across all sites, by dividing the total number of wildings in that type by the total area of that type.
Wildings in cells composed of multiple types were counted in proportion to cell's composition; for example, a cell with one wildling and half its area covered by a given type would tally 0.5 wildling for that type.

For our second analysis, we used a directed acyclical graph (DAG) to diagram causal relationships among the factors we expected to influence the number of wildlings per cell (fig. \@ref(fig:DAG)).
We included an effect of relative elevation on seed rain because neither of our models of seed dispersal capture potential parameter changes induced by uneven terrain [@katulEffectsGentleTopographic2012].   
In the DAG, the unmeasured, proximate causes of wildling number --- establishment likelihood and seed rain, see eq. \@ref(eq:estmod) --- are decendants of variables that we could observe or model.
To estimate the causal influence of ecosystem type on wildlings, the directed acyclical graph dictated that we adjust for site, climate, relative elevation, and relative seed rain in our regression model [@textorRobustCausalInference2016; @mcelreathHauntedDAGCausal2020].
We did not need to adjust for plantation age (with site in the model), but we included it as a covariate anyway because it could reduce the unexplained variance associated with site, and because our assumed causal structure would allow us to interpret its coefficient as an (unconfounded) total effect on wildling number [@westreichTableFallacyPresenting2013].
Climate entered the regression model as mean annual temperature (Bio1) and precipitation of the coldest quarter (Bio19), at 30-arcsecond resolution, from the CHELSA data product [@kargerClimatologiesHighResolution2017].
We chose these two variables because they showed the strongest correlations with Norwegian vegetation zones and sections [@bakkestuenSteplessModelsRegional2008].
Relative elevation was taken with respect to the highest point of the central plantation, from digital elevation models at 1 or 10 m resolution (Norwegian Mapping Authority).
Lastly, plantation surroundings are likely to experience large stochastic variation in annual seed rain, because seed release will concide with different wind conditions from year to year.
Therefore we also included in our model the first order interaction between relative seed rain and plantation age.

```{r DAG, fig.cap="A directed acyclical graph showing the causal relationships that we use as our point of departure for a statistical model of ecosystem effect on wildling establishment. Red variables causally affect both the ecosystem type and wildling establishment, blue variables causally affect only wildling establishment, and grey variables are unobserved. Green arrows show the causal pathway of interest."}
knitr::include_graphics(here('ms', 'figures', 'dagitty-model.svg'))
```

For all species, the number of wildlings per cell showed a high frequency of zeros that was underestimated by the best fitting negative binomial distribution.
Accordingly, we applied zero-inflated, negative binomial (ZINB) generalized linear models.
These regard zeros as the mixed product of a binomial process as well as a (conditional) count process.
We modeled the binomial process as dependent on plantation age and site. 
We expected that younger plantations would exhibit more cells without wildlings than predicted under a constant rate of colonization, because of their infertile juvenile period.
We also expected the frequency of zeros to vary with site, because our field work documented that some land owners had occasionally made efforts to remove wildlings.
Excess zeros that arose in these ways would therefore not bias our estimates of establishment likelihood [@blasco-morenoWhatDoesZero2019].
To summarize, we modeled:

\begin{equation}
wildings_{ij} \sim ZINB(\pi_{ij}, \mu_{ij}, \phi) \\
logit(\pi_{ij}) = PlantationAge_{ij} + Site_{ij} \\
log(\mu_{ij}) =  + \sum\limits_{k=1}^K EcosystemType_{ijk} + Site_{i} + Bio1_{ij} + Bio19_{ij} + RelativeElevation_{ij} + \\
RelativeSeedRain_{ij} + PlantationAge_{ij} + RelativeSeedRain_{ij} \cdot PlantationAge_{ij}  \\
Site_{i} \sim Normal(0, \sigma^2) \\
(\#eq:statmod)
\end{equation}

where $\pi$ is the probability of a zero from the binomial process, while $\mu$ and $\phi$ are the mean and dispersion of the negative binomial distribution, respectively [eq. 11.21 in @zuurZeroTruncatedZeroInflatedModels2009]. 
Subscripts $i$, $j$, and $k$ index sites, cells, and ecosystem types.

We standardized variables (except ecosystem types, which were [0,1]) and fitted models with the glmmTMB package [version 1.0, @brooksGlmmTMBBalancesSpeed2017] in R [version 3.6, @rcoreteamLanguageEnvironmentStatistical2020].
For each species we dropped ecosystem types that were completely free of wildlings (to avoid model convergence issues stemming from complete separation).
We fitted parallel models for our two seed rain variables and selected the model with the better AIC.
To catch problems with our model specification, we looked for deviation from uniformity in quantile-scaled, simulated residuals, using the DHARMa package [version 0.2.7, @hartigDHARMaResidualDiagnostics2020].
We also ran DHARMa's tests for residual over/underdispersion and zero-inflation.
The Norway spruce model showed residual underdispersion, so we replaced the negative binomial distribution with the generalized Poisson distribution, which solved the issue [@brooksStatisticalModelingPatterns2019].
Furthermore, the lodgepole pine model did not converge as specified.
To achieve convergence we dropped site from the binomial submodel, reasoning that its variance was difficult to estimate from only six sites.

Establishment likelihoods for ecosystem types were calculated as predictions from the conditional part of the zero-inflation model (holding covariates at their mean values), scaled by the predicted value for a reference ecosystem type ("forest").
To help summarize patterns in establishment likelihood among ecosystem types, we aggregated ecosystem types by their category (terrestrial or wetland) or structuring process (anthrogenic, natural or no disturbance), as defined in the Nature in Norway system (Appendix [A3]).
We then refitted our (otherwise unchanged) models twice, once with each of these stratifications replacing ecosystem types, and obtained estimates for each category and structuring process.

# Results

The total, unstratified density of wildings in our data set was 28/ha for Sitka/Lutz spruce, 6/ha for Norway spruce, 13/ha for larch, and 34/ha for lodgepole pine.
Observed wildlings densities in different ecosystem types --- excluding ecosystem types covering less than 1 ha --- ranged 1-60/ha for Sitka spruce, 0-49/ha for Norway spruce, 0-52/ha for larch, and 0-219/ha for lodgepole pine.
The "open active landslide" type exceeded these densities around Sitka/Lutz spruce and larch plantations, but was very rare.

For all species, modeled establishment likelihoods varied more than the corresponding wildling densities calculated directly from the data (fig. \@ref(fig:establishment)).
Modeled establishment likelihoods among ecosystem types tallying at least one wildling and covering at least 1 ha varied by up to five orders of magnitude.
Within species, confidence intervals on modeled establishment likelihoods in pairs of ecosystem types frequently overlapped.
Nevertheless, establishment likelihood in many other ecosystem types could be distinguished with a high degree of confidence.
For example, it is not clear whether the establishment likelihood of lodgepole pine was higher in the "coastal heath" or "bare rock" types, but it was clearly much higher in both of these than in "forest".
The pattern of establishment likelihood among ecosystem types was modestly similar between species, with positive rank correlations in four of six species pairs (tab. \@ref(tab:correlation-table)).

Establishment likelihood varied strongly between Sitka/Lutz spruce sites, such that it swamped much of the variation between ecosystem types. 
For example, the model predicted that "arable fields" at the site with the highest overall establishment likelihood and "boreal heath" at the site with the lowest overall establishment likelihood would be colonized at roughly equal rates, even though "boreal heath" is colonized more than 100 times faster on average.
For the other species, especially for Norway spruce and lodgepole pine, there was much less variation in establishment likelihood among sites. 

```{r establishment, fig.cap="Establishment likelihoods among ecosystem types, using 'forest' as the reference level, for four alien conifer species groups. Unfilled points show scaled wildling densities calculated directly from the data. In the case of zero wildlings, these points are placed at the lower limit of the x-axis. Filled points with line intervals show modeled estimates surrounded by their 95 % confidence intervals. Grey point clouds show modeled estimates for individual sites in our data set. Ecosystem types are ordered by a confidence-weighted mean of their percentile-ranked establishment likelihoods within species, such that types with consistently high establishment likelihood across species are at the top of the y-axis. Ecosystem types shown in italic font are wetlands, and ecosystem types shown in bold font are structured by disturbance. Ecosystem types without wildlings and covering less than 0.5 ha across all sites for that species are not displayed. Photos licensed CC BY 4.0 Rune Halvorsen."}
knitr::include_graphics(here('ms', 'figures', 'establishment', 'establishment.svg'))
```

```{r correlation-table}
cortab <- read.csv(here('ms', 'tables', 'spearman-corr.csv'), 
                   row.names = 1, check.names = FALSE)
cortab[lower.tri(cortab, diag = TRUE)] <- ""
cortab <- cortab[-4,-1]
knitr::kable(cortab, 
             caption = "Rank correlations of establishment likelihoods in ecosystem types, between pairs of species.") %>%
  kable_styling(full_width = F)
```

Naturally, the variation in establishment likelihoods shrank when we used category or structuring process to aggregate ecosystem types (fig. \@ref(fig:establishment-hypotheses)).
No more than two orders of magnitude separated establishment likelihoods of these different groupings.
Despite larger sample sizes, many of the groupings were not distinguishable with high confidence.
Sitka/Lutz spruce and Norway spruce established at higher rates in wetland ecosystem types than terrestial ecosystem types, while the opposite was true for larch and lodgepole pine.
Sitka/Lutz spruce and Norway spruce did not colonize disturbance-structured ecosystem types at a higher rate than ecosystem types without disturbance structuring.
For larch and lodgepole pine, structuring by natural distubance and anthropogenic disturbance, respectively, were associated with the highest rates of colonization. 

```{r establishment-hypotheses, fig.cap="Establishment likelihoods among categories of ecosystem types (top) or structuring processes in ecosystem types (bottom), for four alien conifer species groups. Unfilled points show scaled wildling densities calculated directly from the data. Filled points with line intervals show modeled estimates surrounded by their 95 % confidence intervals. The scale of the x-axis differs by species group and the horizontal dotted line marks the reference level for the comparison."}
knitr::include_graphics(here('ms', 'figures', 'establishment-hypotheses.svg'))
```

The direct effect of climate on establishment --- not including its effect mediated by ecosystem types --- was negligible for Sitka/Lutz spruce and Norway spruce, with model estimates near zero (Appendix [A5]). 
For larch there was a negative direct effect of precipitation (Bio19), with a 50 mm increase in coldest-quarter precipitation estimated to decrease establishment six-fold.
From the six sites with lodgepole pine, the model estimated a very strong negative direct effect of annual temperature (Bio01), with a drop of 2.3 ºC decreasing establishment by five orders of magnitude.
It also estimated a strong positive effect of precipitation (Bio19), with a 175 mm increase in coldest-quarter precipitation increasing establishment 500 times.
The direct effect of relative elevation on establishment --- not including its effect mediated by ecosystem types --- was modest and acted in different directions for different species.
For all species, estimates of relative seed rain derived from the WALD model of seed dispersal fit patterns of wildling establishment better than estimates derived from the generalized seed dispersal kernel that we tested.
There was a high degree of confidence that the effect of relative seed rain was positive for all species.
The 95 % confidence intervals around plantation age effects included zero in all cases except in the binomial submodels for Sitka/Lutz spruce and larch (where older plantations showed fewer zeros).
The interaction specified between relative seed rain and plantation age was estimated to be positive for Sitka/Lutz spruce, nonexistent for Norway spruce, and negative for larch and lodgepole pine.

# Discussion

## Ecosystem invasibility

1. Compare results to other assessments conifer recruitment in different ecosystems [@richardsonPineInvasionsSouthern1994; @petersonEcologyManagementSitka1997; @despainDispersalEcologyLodgepole2001; @langdonPinusContortaInvasion2010; @fernandesNaturalRegenerationPinus2016; @taylorDriversPlantInvasion2016; @macekLifeDeathPicea2017; @vanettenInvasionPatternsPinus2020] or to plants more broadly [@chytryHabitatInvasionsAlien2008].
Two obvious management recommendations: 1) for site selection, try to limit plantations to sites where they are surrounded by arable fields. Where it is simultanueosly desireable to alter adjacent nature types (e.g. by grazing), that is also an option. 2) For early detection and rapid response, prioritize monitoring and control interventions in vulnerable ecosystem types like... 5500 km2 mapped for ecosystem types in Norway. Ensuring that plantation surroundings are actively cultivated (e.g. mowed) is an effective way to keep wildlings in check, to the extent that seeds are unlikely to disperse beyond these ecosystem types [@albertLanduseChangeSubalpine2008].

Caution: Beware that Sitka colonization varies enormously from site to site --- much more than other species --- so our ability to predict the magnitude of spread at a specific site is limited. More research is needed to probe this unexplained variance (human action?). But that does not mean we cannot or should not use the estimates to guide management! For example, imagine one site consists entirely of coastal heath and another entirely of arable field. All things being equal, we expect the coastal heath site to be colonized more than 100 times faster than the arable field site. Simultaneusly we should not be surprised if they are colonized at equal rates. 
Possible there are demographic differences between sites for unknown reasons [@taylorDriversPlantInvasion2016]. Shows the continued need for monitoring.

Limitation: inland versus coastal P.contorta sites. Number of wildlings, climate...
It is useful to consider at what spatial scale variability in natural recruitment occurs. This may differ between species. If most variation occurs at the coarsest spatial scale, drivers that also vary at course scales, such as climate, are likely to be comparatively important. If variation between plantations is greater than variation between regions, climate and other coarse-scale factors are probably less important, and drivers that vary at the plantation scale are likely important. If more variation occurs around single plantations than between plantations, site characteristics (e.g. surrounding ecosystem types) are likely the most important governing factors. With only six sites in two distinct groups, differences attributed to climate may have arisen by other mechanisms. Climate may have masked correlated drivers (e.g. sytematic differences in ecosystem composition) 

Advantage: establishment patterns are likely to hold in other countries [@chytryHabitatInvasionsAlien2008]. Because the typification system we used is rule-based and aims to be observer-neutral, land managers in other countries can make educated inferences based on our results, even if a compatible or similar typication system does not exist. For example...

## Accounting for seed rain

2. Accounting for seed rain (and other influences on establishment) amplifies the differences between ecosystem types. Suggests that differential seed rain "averages out" differences in establishment to some extent --- i.e. that, in our sample, vulnerable ecosystems recieved comparatively less seed rain.
For management: do not underestimate the influence of plantation surroundings --- site selection matters! Direct observations of relative abundances misguide intuition about relative vulnerability. For example, the density of Sitka spruce wildlings was about equal in boreal heath and sand dunes, but our estimates suggest that establishment likelihood is actually about 100 times larger in boreal heath.  

In the results from the naive approach, variation in the density of wildlings is partly attributable to differences between species and between ecosystem types, and partly to contingent properties of our sample, such as the spatial distribution of ecosystem types around plantations or the sites' climates).
For example, if a particular ecosystem type tended to occur close to plantations, it would likely show higher wildling density than other ecosystem types even if the species were equally likely to establish there.
Here, seed rain mirrors the concept of propagule pressure; "propagule pressure" often refers to levels of introduction at continental scales, but it also functions as a normalizing quantity (an exposure) for an observed establishment/invasion count [@chytrySeparatingHabitatInvasibility2008].

We can attribute most of the difference between the naive and mod circles in fig. \@ref(fig:establishment) to  

Advantage: compared to tallying analysis, we see that the magnitudes of differences between ecosystem types is much larger. For management the difference between a ten-fold gap and a 1000-fold gap matters --- and may draw more attention to plantation surroundings.

Limitation: relative rather than absolute establishment. Absolute numbers of wildlings matters for management, but estimating absolute magnitude of a seed shadow over the lifetime of the plantation would require knowing lifetime reproductive output, which varies with more than just age (density, site quality, provenance). At the same time, absolute seed rain is always likely to be a highly uncertain quantity, so absolute establishment rates wouldnt have much value anyway. 

Limitation: WALD model of seed rain fits patterns of *establishment* better than ExP, but we cannot strictly test if it was a better descriptor of seed dispersal specifically. Survivorship bias. See different definitions of seed dispersal kernel in Nathan. What processes other than dispersal could give WALD a better fit to establishment than ExP?   
- Negative density dependence, since the WALD model gave a more evenly spread probability distribution. Likely to exist? If it does exist, it would mean that our results overestimate establishment likelihoods near the seed sources and underestimate further away.  


## Patterns of invasibility

3. The fact that disturbance events promote invasion does not map directly to ecosystems characterized by distubance being at higher risk than ecosystems not characterized by disturbance. Or, more broadly, invasion patterns of alien conifers frequently clash with our a priori expectations based on their autecology and general tenets of invasion ecology.
Our disturbance results should not be interpreted to mean that an anthropogenic disturbance event decreases establishment likelihood relative to prior state (e.g. [@vikaneInvasionCallunaHeath2013] shows that burning in coastal heathland increases Sitka spruce establishment), but that alien conifers are not as likely to establish in anthropogenically disturbed ecosystems --- including ecosystems like arable fields --- as a group. An obvious explanation for this pattern is that these kinds of disturbance regimes kill as many or more wildlings than they recruit [@buckleyDisturbanceInvasionReinvasion2007].
We refer to disturbance-structured ecosystems rather than disturbed ecosystems for the above reason.

Ecosystem types structured by anthropogenic disturbance structuring includes areas in which conifer recruitment is almost certainly impossible, like paved surfaces and plowed fields. These deflate the establishment likelihood of this stratum, masking high establishment likelihood in "boreal heaths" or  "strongly altered ground with semi−natural grassland character", for example.  

In some cases, disturbed ecosystems may also be less favorable to establishment because high germination is outweighed by low survival at older life stages. E.g. Gregory 1956 (in Peterson 1997)

Limitation: ecosystems with disturbance of very different frequencies and severities are lumped together in the classification of structuring processes. withing the groups of anthropgenic disturbance and natural disturbance, disturbance frequency and severity varies substantially. 

## Overall

Caution: do not draw conclusions about the overall invasiveness of species from the comparison of establishment likelihoods. The comparison is done for a particular plantation age, under particular climatic conditions, with a particular amount of seed rain --- all of which may vary systematically between species. For example, most lodgepole pine plantations are planted inland but the comparison is under a more oceanic climate, which increases its predicted rate of establishment. Similarly, *Larix* seeds have a higher terminal velocity than *P. abies* and therefore are unlikely to travel as far, but our comparison is at equal seed rain probability density.
More broadly: comparing invasiveness begs the question: "invasive where?" Differences in regional distribution makes it a bit apples and oranges. 
That said, the total wildling density of the species in our sample is congruent with assessments of their invasion potential by the Norwegian impact assessment of alien species, which scored larches one level less invasive than lodgepole pine and Sitka spruce [@Fremmedartslista2018; @sandvikGenericEcologicalImpact2019].      

Advantage: observational study (registering established individuals rather than doing seeding plots) is that seeds experience only natural conditions, and it measures long-term survival --- the quantity of interest for management. For practical reasons, seeding studies observe only the youngest life-stages, and the factors controlling individual success likely differ at later life-stages [@dovciakSeedRainEnvironmental2008]. Therefore, their results are not directly applicable to spread on the longer time-scales of interest to managers.

Limitation: spatial grain. Minimum polygon area NiN, wildling group locations. A good reason to use relatively course observation units (100 m2) and to perform fuzzy coding of ecosystem types. Did our best to represent ecotones in the models and not use polygon boundaries overconfidently. Microsites! (although not a limitation considering applicability to management)

Limitation: wildling establishment varies not only in number and distance but also in growth rate across ecosystem types. Abundant but slow colonization may be of lesser concern than sparse but fast colonization, depending on management objectives.

Caution: our models do not estimate the *total* causal effect of climate or relative elevation on wildlings.
The reason is that each has a causal relationship to wildling number that passes through ecosystem type, which we have adjusted for in our models [@westreichTableFallacyPresenting2013].
Instead, our models estimate the *direct* effects of climate (through establishment likelihood alone) and of relative elevation (through seed rain alone).   
This distinction means that their estimates should not be used guide management directly.
For example, using the climate coefficients (alone) to project risk of wildling establishment across Norway would assume a uniform distribution of ecosystem types everywhere.
Rather, we can interpret their results with respect to specific mechanisms.
Wet and warm conditions...
Locations above plantations...
[@tingstadTemperaturePrecipitationBiotic2015] found that, across Norwegian climatic gradient, effects of climate on Norway spruce recruitement were small (compared to competition). 

Similar methods can be applied to non-conifer plantations that show large site-level variation [e.g. @fernandesWhatDrivesEucalyptus2018].
Wildling spread from plantations is a growing problem [@richardsonConifersInvasiveAliens2004] and is likely to accelerate with recent pushes to increase tree planting worldwide. 

# Appendix

## A1

```{r load sites-table, include=FALSE}
locs <- read_csv(here('output', 'localities-modified.csv')) # see 'wildlings-NiN.Rmd'
locs <- locs %>% 
  select(year.registered, reference, species, 
         locality, utm33.easting, utm33.northing,
         height.source, age, bio01, bio19,
         height.interpolated, age.interpolated) %>% 
  arrange(year.registered, reference, species) %>% 
  select(-year.registered)
# Formating
locs <- locs %>% 
  mutate(species = case_when(
    species == "P.contorta" ~ "Pinus contorta",
    species == "P.sitchensis-lutzii" ~ "Picea sitchensis / lutzii",
    species == "P.abies" ~ "Picea abies",
    species == "Larix" ~ "Larix spp.")) %>% 
  mutate(locality = paste(toupper(substring(locality, 1, 1)), 
                          substring(locality, 2), sep = ""),
             .before = utm33.easting) %>%
  mutate_at(c("utm33.easting","utm33.northing"), round) %>% 
  mutate_at(c("bio01","bio19"), ~ round(./10, digits = 2)) %>% 
  rename(`species group` = species, site = locality, easting = utm33.easting, 
         northing = utm33.northing, height = height.source)
# Footnotes
names(locs)[8] <- paste0(names(locs)[8], footnote_marker_alphabet(1))
names(locs)[9] <- paste0(names(locs)[9], footnote_marker_alphabet(2))
locs <- locs %>% 
  mutate_at(c("height","age"), as.character) %>% 
  mutate(height = case_when(
    height.interpolated ~ paste0(height, footnote_marker_symbol(1)),
    TRUE ~ height)) %>% 
  mutate(age = case_when(
    age.interpolated ~ paste0(age, footnote_marker_symbol(2)),
    TRUE ~ age)) %>% 
  select(-height.interpolated, -age.interpolated)
```

```{r sites-table}
kable(locs, escape = FALSE, align = "lllrrrrrr") %>% 
  kable_styling(bootstrap_options = c("striped", "condensed")) %>% 
  column_spec(2, italic = TRUE) %>%
  footnote(general = "Easting and Northing are given for UTM zone 33N. Height is given in meters and age in years.",
           alphabet = c("mean annual temperature (°C)",
                        "precipitation in coldest quarter (cm)"),
           symbol = c("interpolated as the mean height of other plantations of the same species, inversely weighted by difference in age",
                      "interpolated as the mean age of other plantations of the same species in the same region"))
```

## A2

```{r recreate traits-table, include=FALSE}
traits <- tribble(
  ~`species group`, ~`seed terminal velocity`, ~`dispersal season`, ~references,
  "Larix spp.", 1.0, "Dec - May", "2, 3",
  "Picea abies", 0.58, "Nov - May", "2, 4",
  "Pinus contorta", 0.82, "Sep - Dec", "2",
  "Picea sitchensis / lutzii", 0.94, "Oct - Feb", "1, 2")
```

```{r traits-table}
kable(traits, escape = FALSE) %>% 
  kable_styling(bootstrap_options = c("striped", "condensed")) %>% 
  column_spec(1, italic = TRUE) %>%
  footnote(general = c("1. Harris, A. S. Sitka spruce. in Silvics of North America: 1. Conifers (eds. Burns, R. M. & Honkala, B. H.) vol. 2 513–529 (U.S. Department of Agriculture, Forest Service, 1990).",
                       "2. Sandvik, H. Kunnskapsstatus for spredning og effekter av fremmede bartrær på biologisk mangfold. (2012).",
                       "3. Sullivan, J. Larix decidua. Fire Effects Information System, U.S. Department of Agriculture, Forest Service, Rocky Mountain Research Station,  Fire Sciences Laboratory https://www.fs.fed.us/database/feis/plants/tree/lardec/all.html (1994).",
                       "4. Sullivan, J. Picea abies. Fire Effects Information System, U.S. Department of Agriculture, Forest Service, Rocky Mountain Research Station,  Fire Sciences Laboratory https://www.fs.fed.us/database/feis/plants/tree/picabi/all.html (1994)."), 
            general_title = "References: ")
```

## A3

```{r make types table, include=FALSE}
pairing <- read_csv(here('data','codetype-pairing.csv'))
types <- read_csv(here('data', 'types.csv'))
typestab <- inner_join(types, pairing) %>%
  filter(!grepl(pattern = '.', x = .data$code, fixed = TRUE)) %>% 
  select(type = name, code, category, structuring, `vegetation height` = vegheight)
# Footnotes
names(typestab)[5] <- paste0(names(typestab)[5], footnote_marker_alphabet(1))
```

```{r types table}
options(knitr.kable.NA = '')
kable(typestab, escape = FALSE) %>% 
  collapse_rows(columns = 1) %>% 
  kable_styling(bootstrap_options = c("striped", "condensed")) %>% 
  footnote(alphabet = "approximate vegetation heights (meters) are used only to estimate wind turbulence")
```

## A4

Seed source polygons for which the species of interest only made up a fraction of the plantation composition [e.g. in @olsenKartleggingAvKortdistansespredning2019] were used with their point source density adjusted accordingly.
For example, a plantation identified as composed of Sitka spruce and Norway spruce was assigned a seed source point density half that of a pure Sitka spruce plantation.
Likewise, 'mixed forest' plantations [e.g. in @appelgrenKartleggingAvKortdistansespredning2018] were assigned 0.1 times the seed source point density of a pure plantation.  
Seed source polygons identified as logged [e.g. in @appelgrenKartleggingAvKortdistansespredning2018] were included as seed sources only if we could confirm that they were logged no earlier the decade prior to mapping, using time series of aerial photos.

## A5

```{r, message=FALSE}
msumms <- list.files(here('ms', 'tables'), pattern = "summary", full.names = TRUE) %>% 
  set_names(~ tools::file_path_sans_ext(basename(.))) %>% 
  map(read_csv) %>% 
  map(~ mutate(., Term = if_else(Term == "locality", "site", Term)))
```

```{r}
kable(msumms$`Ps-modelsummary`, align = 'clrrrr', caption = "Picea sitchensis / lutzii. Conditional model is glmmTMB's nbinom2 family with dispersion parameter phi = 0.253.") %>%
  kable_styling() %>% 
  add_header_above(c(" " = 2, "Fixed effects" = 2, "Random effects" = 2)) %>% 
  column_spec(1, bold = T) %>% 
  collapse_rows(columns = 1, valign = "top")

kable(msumms$`Pa-modelsummary`, align = 'clrrrr', caption = "Picea abies. Conditional model is glmmTMB's genpois family with dispersion parameter 2.59.") %>%
  kable_styling() %>% 
  add_header_above(c(" " = 2, "Fixed effects" = 2, "Random effects" = 2)) %>% 
  column_spec(1, bold = T) %>% 
  collapse_rows(columns = 1, valign = "top")

kable(msumms$`L-modelsummary`, align = 'clrrrr', caption = "Larix. Conditional model is glmmTMB's nbinom2 family with dispersion parameter phi = 0.261.") %>%
  kable_styling() %>% 
  add_header_above(c(" " = 2, "Fixed effects" = 2, "Random effects" = 2)) %>% 
  column_spec(1, bold = T) %>% 
  collapse_rows(columns = 1, valign = "top")

kable(msumms$`Pc-modelsummary`, align = 'clrrrr', caption = "Pinus contorta. Conditional model is glmmTMB's nbinom2 family with dispersion parameter phi = 0.0329.") %>%
  kable_styling() %>% 
  add_header_above(c(" " = 2, "Fixed effects" = 2, "Random effects" = 2)) %>% 
  column_spec(1, bold = T) %>% 
  collapse_rows(columns = 1, valign = "top")
```

# References

```{r sessionInfo}
capture.output(sessionInfo(), file="sessionInfo.txt")
```
